---
title: "rhapsodist: method comparison"
author: "Izaskun Mallona"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    self_contained: true
params:
  working_dir:      "/tmp/rhapsodist_simulate"
  sample:           "simulated"
  aligners:         "starsolo,kallisto,alevin"
  has_sbg:          false
  n_expected_cells: 1000
  barcodes_file:    ""
  n_umis_per_cell:  0
  true_mex_dir:     ""
---

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(Matrix)
    library(ggplot2)
    library(ggrepel)
    library(UpSetR)
    library(data.table)
    library(patchwork)
})

wd           <- params$working_dir
samp         <- params$sample
aligners     <- strsplit(params$aligners, ",")[[1]]
has_sbg      <- isTRUE(params$has_sbg) || identical(params$has_sbg, "true")
n_exp        <- as.integer(params$n_expected_cells)
n_umis_truth <- as.integer(params$n_umis_per_cell)
true_mex_dir <- params$true_mex_dir
```

## Data loading

```{r load-sces}
load_sce <- function(aligner, wd, samp) {
    fn <- file.path(wd, aligner, samp, paste0(samp, "_", aligner, "_sce.rds"))
    if (!file.exists(fn)) {
        warning("SCE not found: ", fn)
        return(NULL)
    }
    readRDS(fn)
}

sces <- setNames(lapply(aligners, load_sce, wd = wd, samp = samp), aligners)

if (has_sbg) {
    sbg_fn <- file.path(wd, "sbg", samp, paste0(samp, "_sbg_sce.rds"))
    if (file.exists(sbg_fn)) {
        sces[["sbg"]] <- readRDS(sbg_fn)
    } else {
        warning("SBG SCE not found: ", sbg_fn)
    }
}

sces <- Filter(Negate(is.null), sces)

.rm_zeros <- function(x) {
    x <- x[rowSums(counts(x)) > 0, ]
    x[, colSums(counts(x)) > 0]
}
sces <- lapply(sces, .rm_zeros)

cat("Methods loaded:", paste(names(sces), collapse = ", "), "\n")
sapply(sces, dim) |> `rownames<-`(c("nGenes", "nCells"))
```

## Cell count recovery

```{r cell-counts, fig.width=6, fig.height=4}
counts_df <- data.frame(
    method = names(sces),
    n_cells = sapply(sces, ncol),
    n_genes = sapply(sces, nrow)
)

p <- ggplot(counts_df, aes(x = method, y = n_cells, fill = method)) +
    geom_col(width = 0.6, show.legend = FALSE) +
    geom_text(aes(label = n_cells), vjust = -0.4, size = 3.5) +
    theme_bw() +
    labs(title = "Cells recovered per method", x = NULL, y = "Number of cells") +
    scale_fill_brewer(palette = "Set2")

if (n_exp > 0) {
    p <- p +
        geom_hline(yintercept = n_exp, linetype = "dashed", colour = "firebrick") +
        annotate("text", x = Inf, y = n_exp, label = paste0("expected: ", n_exp),
                 hjust = 1.1, vjust = -0.4, colour = "firebrick", size = 3)
}
p
```

## Ground-truth barcode recovery

```{r ground-truth, eval=nchar(params$barcodes_file) > 0}
truth <- readLines(params$barcodes_file)
cat("Ground-truth cells:", length(truth), "\n\n")

gt_df <- rbindlist(lapply(names(sces), function(m) {
    cb <- colnames(sces[[m]])
    data.frame(
        method    = m,
        n_cells   = length(cb),
        n_true    = sum(cb %in% truth),
        n_false   = sum(!(cb %in% truth)),
        recall    = mean(cb %in% truth),
        precision = sum(cb %in% truth) / length(truth)
    )
}))
knitr::kable(gt_df, digits = 3,
    caption = paste("Recall = fraction of recovered cells in ground truth;",
                    "Precision = fraction of ground-truth cells recovered"))
```

```{r gt-plot, eval=nchar(params$barcodes_file) > 0, fig.width=7, fig.height=4}
gt_long <- melt(as.data.table(gt_df[, c("method", "n_true", "n_false")]),
                id.vars = "method", variable.name = "type", value.name = "count")
gt_long[, type := ifelse(type == "n_true", "True positive", "False positive")]

ggplot(gt_long, aes(x = method, y = count, fill = type)) +
    geom_col(width = 0.6) +
    scale_fill_manual(values = c("True positive" = "#66c2a5",
                                  "False positive" = "#fc8d62")) +
    theme_bw() +
    labs(title = "Cell barcode recovery vs ground truth",
         x = NULL, y = "Number of cells", fill = NULL)
```

## Barcode overlaps

```{r upset, fig.width=8, fig.height=5}
bc_lists <- lapply(sces, colnames)
upset(fromList(bc_lists), order.by = "freq", text.scale = 1.4,
      point.size = 3, line.size = 0.8, mb.ratio = c(0.6, 0.4))
```

## Deviation from simulated truth

```{r truth-load, eval=nchar(true_mex_dir) > 0}
.find_mex <- function(dir, base) {
    gz <- file.path(dir, paste0(base, '.gz'))
    if (file.exists(gz)) return(gz)
    file.path(dir, base)
}
truth_counts   <- readMM(.find_mex(true_mex_dir, 'matrix.mtx'))
truth_barcodes <- readLines(.find_mex(true_mex_dir, 'barcodes.tsv'))
truth_features <- read.table(.find_mex(true_mex_dir, 'features.tsv'),
                              sep = '\t', header = FALSE)
rownames(truth_counts) <- truth_features$V1
colnames(truth_counts) <- truth_barcodes
truth_pb <- rowSums(truth_counts)
```

```{r truth-heatmap, eval=nchar(true_mex_dir) > 0, fig.width=5, fig.height=4}
## Build a pseudo-bulk matrix that includes truth as an additional column,
## then compute pairwise mean absolute relative difference (MARD) between
## all methods and truth.  MARD = mean(|x-y| / ((x+y)/2)).
shared_genes <- Reduce(intersect, c(lapply(sces, rownames), list(names(truth_pb))))

pb_with_truth <- do.call(cbind, c(
    lapply(names(sces), function(m) rowSums(counts(sces[[m]]))[shared_genes]),
    list(truth = truth_pb[shared_genes])
))
colnames(pb_with_truth) <- c(names(sces), "truth")

mard_mat <- outer(
    colnames(pb_with_truth), colnames(pb_with_truth),
    Vectorize(function(a, b) {
        x <- pb_with_truth[, a]
        y <- pb_with_truth[, b]
        denom <- (x + y) / 2
        mean(abs(x - y)[denom > 0] / denom[denom > 0], na.rm = TRUE)
    })
)
dimnames(mard_mat) <- list(colnames(pb_with_truth), colnames(pb_with_truth))

mard_long <- as.data.frame(as.table(mard_mat))
names(mard_long) <- c("method1", "method2", "mard")

ggplot(mard_long, aes(method1, method2, fill = mard * 100)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = sprintf("%.2f%%", mard * 100)), size = 3.2) +
    scale_fill_distiller(palette = "RdYlGn", direction = -1, limits = c(0, NA)) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
    labs(title = "Pairwise mean absolute relative difference",
         subtitle = "Including simulated truth; lower is better",
         x = NULL, y = NULL, fill = "MARD (%)")
```

```{r truth-summary, eval=nchar(true_mex_dir) > 0}
truth_df <- rbindlist(lapply(names(sces), function(m) {
    pb <- rowSums(counts(sces[[m]]))[shared_genes]
    rel_err <- (pb - truth_pb[shared_genes]) / truth_pb[shared_genes]
    data.table(method = m, gene = shared_genes, rel_err = rel_err)
}))
truth_sum <- truth_df[, .(
    mean_rel_err   = round(mean(rel_err), 4),
    median_rel_err = round(median(rel_err), 4),
    rmse           = round(sqrt(mean(rel_err^2)), 4)
), by = method]
knitr::kable(truth_sum,
    caption = "Per-method deviation from simulated truth (relative error per gene)")
```

## Accuracy vs speed

```{r accuracy-vs-speed, eval=nchar(true_mex_dir) > 0, fig.width=6, fig.height=4, message=FALSE, warning=FALSE}
bm_dir   <- file.path(wd, "benchmarks")
bm_files <- list.files(bm_dir, pattern = "\\.txt$", full.names = TRUE)

if (length(bm_files) > 0) {
    bm <- rbindlist(Filter(Negate(is.null), lapply(bm_files, function(fn) {
        tryCatch({
            dt <- fread(fn, sep = "\t")
            dt[, step := gsub("\\.txt$", "", basename(fn))]
            dt
        }, error = function(e) NULL)
    })), fill = TRUE)

    classify_pipeline <- function(step) {
        if (grepl("^starsolo|^star|r_sce_generation_starsolo|starsolo_descriptive", step)) return("starsolo")
        if (grepl("^kallisto|^bustools|r_sce_generation_kallisto|kallisto_descriptive", step)) return("kallisto")
        if (grepl("^alevin|r_sce_generation_alevin|alevin_descriptive", step)) return("alevin")
        if (grepl("^sbg|r_sce_generation_sbg|sbg_descriptive", step)) return("sbg")
        return(NA_character_)
    }
    bm[, pipeline := sapply(step, classify_pipeline)]
    bm[, minutes  := s / 60]

    pipeline_sum <- bm[!is.na(pipeline),
                       .(total_min = sum(minutes)), by = pipeline]

    acc <- truth_sum[, .(pipeline = method, rmse)]

    cmp <- merge(pipeline_sum, acc, by = "pipeline")

    ggplot(cmp, aes(x = total_min, y = rmse, colour = pipeline, label = pipeline)) +
        geom_point(size = 4) +
        ggrepel::geom_text_repel(size = 3.5, show.legend = FALSE) +
        scale_colour_brewer(palette = "Set2") +
        theme_bw() +
        labs(title = "Accuracy vs speed",
             subtitle = "RMSE of relative error vs simulated truth",
             x = "Total pipeline time (minutes)",
             y = "RMSE (relative error vs truth)",
             colour = NULL)
} else {
    cat("No benchmark files found; skipping.\n")
}
```

## Per-cell cross-aligner correlation

For every pair of aligners, cells shared by both (matched by cell barcode) are
correlated (Pearson on log1p counts) across the shared gene set. The plot shows
the distribution of these per-cell correlations across the shared population.

```{r per-cell-cor, fig.width=8, fig.height=4.5}
col_pearson <- function(A, B) {
    Am <- colMeans(A); Bm <- colMeans(B)
    Ac <- sweep(A, 2, Am, "-"); Bc <- sweep(B, 2, Bm, "-")
    num <- colSums(Ac * Bc)
    denom <- sqrt(colSums(Ac^2) * colSums(Bc^2))
    ifelse(denom > 0, num / denom, NA_real_)
}

## sbg rownames are gene symbols, while starsolo/kallisto/alevin carry
## Ensembl IDs as rownames and symbols in rowData. Match on the best
## available symbol key and subset counts by integer position, because
## HDF5-backed assays do not always inherit updated dimnames.
gene_key <- function(sce) {
    rd <- SummarizedExperiment::rowData(sce)
    for (col in c("Symbol", "symbol", "gene_name", "gene_symbol")) {
        if (col %in% colnames(rd)) {
            v <- as.character(rd[[col]])
            if (length(v) && all(nzchar(v))) return(v)
        }
    }
    best_col <- NA_character_
    best_u <- 1L
    for (col in colnames(rd)) {
        v <- as.character(rd[[col]])
        u <- length(unique(v))
        if (u > best_u) { best_u <- u; best_col <- col }
    }
    if (!is.na(best_col)) return(as.character(rd[[best_col]]))
    rn <- rownames(sce)
    if (!is.null(rn) && length(rn) && all(nzchar(rn))) return(rn)
    as.character(seq_len(nrow(sce)))
}

gene_keys <- lapply(sces, gene_key)
pair_combos <- if (length(sces) >= 2) combn(names(sces), 2, simplify = FALSE) else list()
max_cells_per_cell_cor <- 500L

set.seed(1)
per_cell_cor <- rbindlist(lapply(pair_combos, function(pair) {
    a <- pair[1]; b <- pair[2]
    shared_cells <- intersect(colnames(sces[[a]]), colnames(sces[[b]]))
    key_a <- gene_keys[[a]]; key_b <- gene_keys[[b]]
    shared_genes <- intersect(key_a, key_b)
    if (length(shared_cells) < 2 || length(shared_genes) < 2) return(NULL)
    ix_a <- match(shared_genes, key_a)
    ix_b <- match(shared_genes, key_b)
    if (length(shared_cells) > max_cells_per_cell_cor) {
        shared_cells <- sort(sample(shared_cells,
                                    max_cells_per_cell_cor,
                                    replace = FALSE))
    }
    A <- log1p(as.matrix(counts(sces[[a]])[ix_a, shared_cells, drop = FALSE]))
    B <- log1p(as.matrix(counts(sces[[b]])[ix_b, shared_cells, drop = FALSE]))
    data.table(pair = paste(a, "vs", b),
               cell = shared_cells,
               correlation = col_pearson(A, B))
}), fill = TRUE)

if (nrow(per_cell_cor) == 0) {
    cat("Fewer than two aligners with shared cells/genes; skipping.\n")
} else {
    valid <- per_cell_cor[!is.na(correlation)]
    ggplot(valid, aes(x = correlation, colour = pair, fill = pair)) +
        geom_density(alpha = 0.25, linewidth = 0.7) +
        scale_colour_brewer(palette = "Dark2") +
        scale_fill_brewer(palette = "Dark2") +
        theme_bw() +
        labs(title = "Per-cell cross-aligner expression correlation",
             subtitle = sprintf(
                 "Pearson on log1p counts, shared genes (symbol-matched), up to %d cells per pair",
                 max_cells_per_cell_cor),
             x = "Correlation", y = "Density", colour = NULL, fill = NULL)
}
```

```{r per-cell-cor-summary}
if (exists("per_cell_cor") && nrow(per_cell_cor) > 0) {
    per_cell_cor_summary <- per_cell_cor[!is.na(correlation),
        .(n_cells = .N,
          median  = round(median(correlation), 3),
          mean    = round(mean(correlation), 3),
          q25     = round(quantile(correlation, 0.25), 3),
          q75     = round(quantile(correlation, 0.75), 3)),
        by = pair]
    knitr::kable(per_cell_cor_summary,
                 caption = "Per-cell cross-aligner correlation summary")
}
```

## Per-cell total-count distribution

```{r umi-dist, fig.width=8, fig.height=4}
umi_df <- rbindlist(lapply(names(sces), function(m) {
    data.table(method = m, source = "recovered", total_counts = colSums(counts(sces[[m]])))
}))

## overlay simulated truth when available
has_truth_counts <- nchar(true_mex_dir) > 0 && exists("truth_counts")
if (has_truth_counts) {
    truth_totals <- colSums(truth_counts)
    umi_df <- rbind(
        umi_df,
        data.table(method = "simulated truth", source = "simulated", total_counts = truth_totals)
    )
}

## fixed colour for truth; Set2 for recovered methods
method_levels <- c(names(sces), if (has_truth_counts) "simulated truth")
set2_cols     <- RColorBrewer::brewer.pal(max(3, length(names(sces))), "Set2")[seq_along(names(sces))]
colour_map    <- setNames(c(set2_cols, if (has_truth_counts) "black"), method_levels)

ggplot(umi_df, aes(x = log10(total_counts + 1),
                   colour = method, linetype = source)) +
    geom_density(linewidth = 0.8) +
    scale_colour_manual(values = colour_map) +
    scale_linetype_manual(values = c(simulated = "dashed", recovered = "solid")) +
    theme_bw() +
    labs(title = "Per-cell total-count distribution",
         subtitle = "recovered methods (solid) vs simulated truth (dashed black)",
         x = "log10(total counts + 1)", y = "Density",
         colour = NULL, linetype = NULL)
```
