---
title: "rhapsodist: computational benchmarks"
author: "Izaskun Mallona"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    self_contained: true
params:
  working_dir:      "/tmp/rhapsodist_simulate"
  n_expected_cells: 1000
---

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
    library(data.table)
    library(ggplot2)
    library(patchwork)
    library(ggrepel)
})

wd    <- params$working_dir
n_exp <- as.integer(params$n_expected_cells)

src_dir <- file.path(dirname(knitr::current_input()), "..")
source(file.path(src_dir, "benchmark_stages.R"))
stage_table <- read_stage_table(file.path(src_dir, "..", "data", "benchmark_stages.tsv"))
cwl_stage_table <- read.delim(file.path(src_dir, "..", "data", "sbg_cwl_stages.tsv"),
                              stringsAsFactors = FALSE)
```

## Machine and run settings

```{r run-info}
run_info_file <- file.path(wd, "run_info.tsv")
if (file.exists(run_info_file)) {
    knitr::kable(fread(run_info_file, sep = "\t"))
} else {
    cat("run_info.tsv not found in", wd, "\n")
}
```

## Load benchmark data

`threads` is filled when Snakemake runs with `--benchmark-extended`, which the
`rhapsodist` command passes.

```{r load-benchmarks}
bm_dir   <- file.path(wd, "benchmarks")
bm_files <- list.files(bm_dir, pattern = "\\.txt$", full.names = TRUE)
## this report's own benchmark is written after it renders; one from an earlier run is left out
bm_files <- bm_files[basename(bm_files) != "benchmarks_report.txt"]
if (length(bm_files) == 0) stop("No benchmark files found in ", bm_dir)

read_bm <- function(fn) {
    tryCatch({
        dt <- fread(fn, sep = "\t")
        dt[, file := basename(fn)]
        dt
    }, error = function(e) NULL)
}

bm <- rbindlist(Filter(Negate(is.null), lapply(bm_files, read_bm)), fill = TRUE)
bm[, stem := gsub("\\.txt$", "", file)]
classified <- classify_benchmarks(bm$stem, stage_table)
bm[, c("step", "pipeline", "aligners", "stage") :=
       classified[, c("step", "pipeline", "aligners", "stage")]]
bm[, minutes  := s / 60]
if (!"threads" %in% names(bm)) bm[, threads := NA_integer_]

cat("Benchmark records loaded:", nrow(bm), "\n")
knitr::kable(
    bm[order(pipeline, step), .(step, pipeline, stage,
                                 seconds = round(s, 1),
                                 minutes = round(minutes, 2),
                                 max_rss_MB = round(max_rss, 1),
                                 threads)][1:min(nrow(bm), 30)],
    caption = "First 30 benchmark entries"
)
```

## Runtime per step

```{r runtime-per-step, fig.width=10, fig.height=5}
bm_steps <- bm[!(stage == "reporting" & pipeline == "shared")]

ggplot(bm_steps, aes(x = reorder(step, minutes), y = minutes, fill = pipeline)) +
    geom_col(width = 0.7) +
    coord_flip() +
    scale_fill_brewer(palette = "Set2", na.value = "grey70") +
    theme_bw(base_size = 11) +
    labs(title = "Wall-clock time per workflow step",
         x = NULL, y = "Time (minutes)", fill = "Pipeline")
```

## Total runtime per pipeline

Only steps assigned to a single pipeline (not "shared") are summed.

```{r runtime-per-pipeline, fig.width=7, fig.height=4}
pipeline_time <- bm_steps[, .(minutes = max(minutes)), by = .(step, pipeline)]
pipeline_sum  <- pipeline_time[, .(total_min = sum(minutes)), by = pipeline]
pipeline_sum[, pipeline := factor(pipeline,
    levels = c("shared", "starsolo", "kallisto", "alevin", "sbg"))]

ggplot(pipeline_sum, aes(x = pipeline, y = total_min, fill = pipeline)) +
    geom_col(width = 0.6, show.legend = FALSE) +
    geom_text(aes(label = round(total_min, 1)), vjust = -0.4, size = 3.5) +
    scale_fill_brewer(palette = "Set2", na.value = "grey70") +
    theme_bw() +
    labs(title = "Total wall-clock time per pipeline",
         x = NULL, y = "Time (minutes)")
```

## Runtime per stage

Seconds per aligner and stage, from `workflow/data/benchmark_stages.tsv`. A step
used by several aligners, such as the barcode standardization, is counted for
each of them. Download, install, reference and simulation steps run once and
are left out. When the BD CWL run wrote its step times, the CWL rule is split
by step (`workflow/data/sbg_cwl_stages.tsv`); its first step,
`QualCLAlign_RNA`, parses barcodes and aligns, so it holds work the other
aligners do in preprocessing.

```{r runtime-per-stage, fig.width=8, fig.height=4}
aligners_run <- intersect(BENCHMARK_ALIGNERS, unique(bm$pipeline))
cwl_files <- list.files(bm_dir, pattern = "^sbg_cwl_steps_.*\\.tsv$", full.names = TRUE)
cwl_steps <- rbindlist(lapply(cwl_files, function(fn) {
    steps <- fread(fn, sep = "\t")
    steps[, sample := sub("^sbg_cwl_steps_(.*)\\.tsv$", "\\1", basename(fn))]
    steps[!is.na(seconds), .(sample, step, seconds)]
}))

by_stage <- as.data.table(stage_times(
    as.data.frame(bm[, .(step, aligners, stage)]), bm$s, aligners_run,
    if (nrow(cwl_steps) > 0) as.data.frame(cwl_steps) else NULL, cwl_stage_table))
by_stage <- by_stage[stage %in% c(COMPARED_STAGES, "CWL overhead", "other modalities")]
by_stage[, stage := factor(stage, levels = c(COMPARED_STAGES, "other modalities", "CWL overhead"))]
by_stage[, pipeline := factor(pipeline, levels = BENCHMARK_ALIGNERS)]

ggplot(by_stage, aes(x = pipeline, y = seconds / 60, fill = stage)) +
    geom_col(width = 0.6) +
    scale_fill_brewer(palette = "Set2", drop = TRUE) +
    theme_bw() +
    labs(title = "Wall-clock time per stage",
         x = NULL, y = "Time (minutes)", fill = "Stage")

knitr::kable(
    dcast(by_stage, stage ~ pipeline, value.var = "seconds",
          fun.aggregate = function(x) round(sum(x) / 60, 2)),
    caption = "Minutes per stage and aligner"
)
```

## Memory usage (max RSS)

```{r memory, fig.width=10, fig.height=5}
bm_mem <- bm_steps[!is.na(max_rss) & max_rss > 0]
if (nrow(bm_mem) > 0) {
    ggplot(bm_mem, aes(x = reorder(step, max_rss), y = max_rss / 1024,
                       fill = pipeline)) +
        geom_col(width = 0.7) +
        coord_flip() +
        scale_fill_brewer(palette = "Set2", na.value = "grey70") +
        theme_bw(base_size = 11) +
        labs(title = "Peak memory (max RSS) per step",
             x = NULL, y = "Memory (GB)", fill = "Pipeline")
} else {
    cat("No memory data available.\n")
}
```

## Speed vs cell recovery

```{r recovery-vs-time, fig.width=6, fig.height=4, eval=n_exp > 0}
pipelines <- c("starsolo", "kallisto", "alevin", "sbg")
pl_dirs <- Filter(dir.exists, file.path(wd, pipelines))
sample_names <- if (length(pl_dirs) > 0) basename(list.dirs(pl_dirs[1], recursive = FALSE)) else character(0)

recovery <- rbindlist(lapply(pipelines, function(pl) {
    rbindlist(lapply(sample_names, function(samp) {
        fn  <- file.path(wd, pl, samp, paste0(samp, "_", pl, "_sce.rds"))
        sce <- tryCatch(readRDS(fn), error = function(e) NULL)
        if (is.null(sce)) return(NULL)
        data.table(pipeline = pl, sample = samp, n_cells = ncol(sce))
    }), fill = TRUE)
}), fill = TRUE)

if (nrow(recovery) > 0) {
    cmp <- merge(
        recovery[, .(n_cells = mean(n_cells)), by = pipeline],
        pipeline_sum, by = "pipeline", all.x = TRUE)
    cmp[, recovery_pct := n_cells / n_exp * 100]

    ggplot(cmp, aes(x = total_min, y = recovery_pct, colour = pipeline,
                    label = pipeline)) +
        geom_point(size = 4) +
        geom_text_repel(size = 3.5, show.legend = FALSE) +
        geom_hline(yintercept = 100, linetype = "dashed", colour = "grey50") +
        scale_colour_brewer(palette = "Set2") +
        theme_bw() +
        labs(title = "Speed vs cell recovery",
             subtitle = paste0("Expected cells: ", n_exp),
             x = "Total pipeline time (minutes)",
             y = paste0("Cells recovered (% of ", n_exp, ")"),
             colour = NULL)
} else {
    cat("No SCE files found for recovery plot.\n")
}
```

## Full benchmark table

```{r full-table}
knitr::kable(
    bm[order(pipeline, step),
       .(step, pipeline, stage,
         seconds    = round(s, 1),
         minutes    = round(minutes, 2),
         max_rss_MB = round(max_rss, 1),
         mean_load  = round(mean_load, 2),
         threads)],
    caption = "Complete benchmark statistics for all workflow steps"
)
```
