---
title: "rhapsodist: biology report"
author: "Izaskun Mallona"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    self_contained: true
params:
  working_dir: "/tmp/rhapsodist_simulate"
  sample: "simulated"
  aligners: "starsolo,kallisto,alevin"
  has_sbg: false
  markers_file: ""
  guide_csv: ""
  use_case: "sendoel"
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
    error = TRUE,
    echo = TRUE,
    cache = FALSE,
    dev = "png",
    dev.args = list(type = "cairo-png"),
    warning = FALSE,
    message = FALSE
)

suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(Seurat)
    library(scuttle)
    library(ggplot2)
    library(patchwork)
    library(data.table)
    library(Matrix)
    library(viridis)
    library(UpSetR)
    library(clue)
    library(mclust)
    library(ggrepel)
})

source(file.path(dirname(knitr::current_input()), "..", "ggtheme.R"))

wd = params$working_dir
samp = params$sample
aligners = strsplit(params$aligners, ",")[[1]]
has_sbg = isTRUE(params$has_sbg) || identical(params$has_sbg, "true")
use_case = if (nzchar(params$use_case)) params$use_case else "sendoel"
stopifnot(use_case %in% c("sendoel", "hela"))
is_hela = identical(use_case, "hela")

if (!is_hela) {
    if (!nzchar(params$markers_file) || !file.exists(params$markers_file)) {
        stop("markers_file not set or not found: '", params$markers_file,
             "'. Set biology_markers_file in the config.")
    }
    marker_meta = read.table(params$markers_file, header = TRUE, sep = "\t",
                             stringsAsFactors = FALSE)
    markers = marker_meta$marker
    stopifnot(!anyDuplicated(marker_meta$marker))
} else {
    marker_meta = data.frame(marker = character(0), cell_type = character(0),
                             stringsAsFactors = FALSE)
    markers = character(0)
}

if (has_sbg) aligners = c(aligners, "sbg")

pipeline_colours = setNames(
    rep(cbbPalette, length.out = length(aligners)),
    aligners)

prefix_overrides = c(starsolo = "s", kallisto = "k", alevin = "a",
                     sbg = "b", bd = "r")
pipe_prefix = function(pipe) {
    if (pipe %in% names(prefix_overrides)) unname(prefix_overrides[pipe])
    else substr(pipe, 1, 1)
}
stopifnot(!anyDuplicated(vapply(aligners, pipe_prefix, character(1))))
```

## Data loading

Each pipeline produces a SingleCellExperiment saved as an RDS file. Gene symbols
sit in rowData under different column names depending on the aligner (starsolo:
`name`, kallisto: `SYMBOL`, alevin: Ensembl IDs as rownames with no symbol
column). SCE rownames are renamed to gene symbols before building Seurat objects
so that marker lookups work the same way across pipelines.

```{r load-sces}
load_sce = function(aligner, wd, samp) {
    fn = file.path(wd, aligner, samp, paste0(samp, "_", aligner, "_sce.rds"))
    if (!file.exists(fn)) {
        warning("SCE not found: ", fn)
        return(NULL)
    }
    readRDS(fn)
}

sces = setNames(lapply(aligners, load_sce, wd = wd, samp = samp), aligners)
sces = Filter(Negate(is.null), sces)

rm_zeros = function(sce) {
    sce = sce[rowSums(counts(sce)) > 0, ]
    sce[, colSums(counts(sce)) > 0]
}
sces = lapply(sces, rm_zeros)

symbol_rownames = function(sce) {
    rd = rowData(sce)
    sym = rownames(sce)
    for (col in c("name", "SYMBOL", "gene_name")) {
        if (col %in% colnames(rd)) {
            sym = as.character(rd[[col]])
            break
        }
    }
    rownames(sce) = make.unique(sym)
    sce
}

sces = lapply(sces, symbol_rownames)

cat("Pipelines loaded:", paste(names(sces), collapse = ", "), "\n")
sapply(sces, dim) |> `rownames<-`(c("n_genes", "n_cells"))
```

## QC summary

Cell and library quality metrics per pipeline. Differences in cell count come
from each aligner's barcode calling approach. Median UMI and genes detected show
how many counts each aligner assigns per cell.

```{r qc-table}
qc_df = rbindlist(lapply(names(sces), function(m) {
    sce = addPerCellQCMetrics(sces[[m]],
        subsets = list(mito = grep("^mt-|^MT-", rownames(sces[[m]]), value = TRUE)))
    data.table(
        pipeline = m,
        n_cells = ncol(sce),
        n_genes_total = nrow(sce),
        median_umi = round(median(sce$sum)),
        median_genes = round(median(sce$detected)),
        median_pct_mito = round(median(sce$subsets_mito_percent), 2)
    )
}))

saveRDS(qc_df, file.path(wd, paste0(samp, "_biology_qc.rds")))
knitr::kable(qc_df, caption = "Per-pipeline QC summary.")
```

Cells per pipeline, median UMI per cell, and median genes detected per cell.
Bars allow quick visual comparison; exact values are in the table above.

```{r qc-barplots, fig.width=ng_fig_size(3, 1)$w, fig.height=ng_fig_size(3, 1)$h}
qc_long = melt(
    qc_df[, .(pipeline, n_cells, median_umi, median_genes)],
    id.vars = "pipeline", variable.name = "metric", value.name = "value"
)

ggplot(qc_long, aes(x = pipeline, y = value, fill = pipeline)) +
    geom_col() +
    geom_text(aes(label = value), vjust = -0.4, size = 2.5) +
    facet_wrap(~metric, scales = "free_y") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    scale_fill_manual(values = pipeline_colours) +
    theme_ng() +
    theme(aspect.ratio = NULL, legend.position = "none") +
    labs(x = NULL, y = NULL)
```

Per-cell UMI and gene count distributions across pipelines. A shift to the right
means more counts per cell. Multimodal distributions may come from doublets or
cell populations of very different sizes.

```{r umi-gene-density, fig.width=ng_fig_size(2, 1)$w, fig.height=ng_fig_size(2, 1)$h}
umi_df = rbindlist(lapply(names(sces), function(m) {
    data.table(pipeline = m, umi = colSums(counts(sces[[m]])))
}))

gene_df = rbindlist(lapply(names(sces), function(m) {
    data.table(pipeline = m, genes = colSums(counts(sces[[m]]) > 0))
}))

p_umi = ggplot(umi_df, aes(x = log10(umi + 1), colour = pipeline)) +
    geom_density(linewidth = 0.6) +
    scale_colour_manual(values = pipeline_colours) +
    theme_ng() +
    theme(aspect.ratio = NULL) +
    labs(x = "log10(UMI + 1)", y = "density", colour = NULL)

p_gene = ggplot(gene_df, aes(x = log10(genes + 1), colour = pipeline)) +
    geom_density(linewidth = 0.6) +
    scale_colour_manual(values = pipeline_colours) +
    theme_ng() +
    theme(aspect.ratio = NULL) +
    labs(x = "log10(genes + 1)", y = "density", colour = NULL)

p_umi + p_gene + plot_layout(guides = "collect")
```

## UMAP {.tabset}

For each pipeline: 2000 highly variable genes by VST, 30 PCs, Louvain clustering
at resolution 0.5, UMAP on the top 20 PCs. Feature plots show log-normalised
expression; cells with zero counts are shown in grey. All panels below show the
pipelines side by side on a common layout so cluster structure and composition
can be compared directly. Cluster numbers are prefixed by pipeline
(alevin: a_N, starsolo: s_N, kallisto: k_N) so labels stay unambiguous across
panels and in the Hungarian confusion matrices below.

```{r build-seurat-objects}
make_seurat = function(sce, fallback_project) {
    mat = as(counts(sce), "CsparseMatrix")
    proj = mainExpName(sce)
    if (is.null(proj) || length(proj) == 0 || !nzchar(proj)) proj = fallback_project
    so = CreateSeuratObject(counts = mat, project = proj,
                            min.cells = 0, min.features = 0)
    so = NormalizeData(so, normalization.method = "LogNormalize",
                       scale.factor = 1e4, verbose = FALSE)
    so = FindVariableFeatures(so, selection.method = "vst", nfeatures = 2000,
                              verbose = FALSE)
    so = ScaleData(so, verbose = FALSE)
    so = RunPCA(so, features = VariableFeatures(so), npcs = 30, verbose = FALSE)
    so = FindNeighbors(so, dims = 1:20, verbose = FALSE)
    so = FindClusters(so, resolution = 0.5, verbose = FALSE)
    so = RunUMAP(so, dims = 1:20, verbose = FALSE)
    so
}

seurat_objects = list()
for (pipe in names(sces)) {
    cache_fn = file.path(wd, paste0(samp, "_biology_", pipe, "_seurat.rds"))
    built = tryCatch({
        if (file.exists(cache_fn)) {
            cat(pipe, ": loading cached Seurat object\n")
            readRDS(cache_fn)
        } else {
            cat(pipe, ": building Seurat object\n")
            so = make_seurat(sces[[pipe]], fallback_project = paste(samp, pipe, sep = "_"))
            saveRDS(so, cache_fn)
            so
        }
    }, error = function(e) {
        warning(sprintf("Seurat build failed for %s: %s", pipe, conditionMessage(e)))
        NULL
    })
    if (!is.null(built)) seurat_objects[[pipe]] = built
}

cat("Seurat objects available for:",
    paste(names(seurat_objects), collapse = ", "), "\n")

rm(sces)
gc()
```

```{r cell-cycle-scoring}
## Seurat ships Tirosh S and G2M gene lists as human symbols. For mouse we
## Title-case them. Pick the organism by overlap with the first Seurat object's
## rownames so sendoel and hela runs alike get cell cycle phase annotations.
s_genes_hs = Seurat::cc.genes.updated.2019$s.genes
g2m_genes_hs = Seurat::cc.genes.updated.2019$g2m.genes
s_genes_mm = stringr::str_to_title(s_genes_hs)
g2m_genes_mm = stringr::str_to_title(g2m_genes_hs)

pick_cc_genes = function(so) {
    rn = rownames(so)
    hs_hits = length(intersect(rn, c(s_genes_hs, g2m_genes_hs)))
    mm_hits = length(intersect(rn, c(s_genes_mm, g2m_genes_mm)))
    if (hs_hits == 0 && mm_hits == 0) return(NULL)
    if (hs_hits >= mm_hits) list(s = s_genes_hs, g2m = g2m_genes_hs, organism = "human")
    else list(s = s_genes_mm, g2m = g2m_genes_mm, organism = "mouse")
}

cc_phase_rows = list()
cc_score_rows = list()
for (pipe in names(seurat_objects)) {
    so = seurat_objects[[pipe]]
    cc = pick_cc_genes(so)
    if (is.null(cc)) {
        warning(pipe, ": no cell-cycle marker overlap; skipping CellCycleScoring")
        so$Phase = NA_character_
        so$S.Score = NA_real_
        so$G2M.Score = NA_real_
    } else {
        so = tryCatch(
            Seurat::CellCycleScoring(so, s.features = cc$s,
                                     g2m.features = cc$g2m, set.ident = FALSE),
            error = function(e) {
                warning(pipe, ": CellCycleScoring failed: ", conditionMessage(e))
                so$Phase = NA_character_
                so$S.Score = NA_real_
                so$G2M.Score = NA_real_
                so
            })
        cat(pipe, ": cell-cycle scored using", cc$organism, "markers\n")
    }
    seurat_objects[[pipe]] = so

    ## refresh the seurat cache so downstream consumers see the cell cycle phase
    cache_fn = file.path(wd, paste0(samp, "_biology_", pipe, "_seurat.rds"))
    saveRDS(so, cache_fn)

    cc_phase_rows[[pipe]] = data.table(pipeline = pipe,
                                       phase = so$Phase)[!is.na(phase),
                                                         .N, by = .(pipeline, phase)]
    cc_score_rows[[pipe]] = data.table(pipeline = pipe,
                                       barcode = colnames(so),
                                       S.Score = so$S.Score,
                                       G2M.Score = so$G2M.Score,
                                       Phase = so$Phase)
}
cc_phase_dt = rbindlist(cc_phase_rows)
cc_score_dt = rbindlist(cc_score_rows)
fwrite(cc_phase_dt,
       file.path(wd, paste0(samp, "_biology_cell_cycle_phase_counts.csv")))
fwrite(cc_score_dt,
       file.path(wd, paste0(samp, "_biology_cell_cycle_scores.csv")))
saveRDS(cc_score_dt, file.path(wd, paste0(samp, "_biology_cell_cycle.rds")))
```

```{r cluster-prefixes-and-guides}
## Always: set per-pipe cluster_prefixed and load guides (if any). Marker-based
## celltype assignment is kept in the next chunk so hela runs can skip it.
for (pipe in names(seurat_objects)) {
    so = seurat_objects[[pipe]]
    so$cluster_prefixed = paste0(pipe_prefix(pipe), "_", as.character(Idents(so)))
    seurat_objects[[pipe]] = so
}

guide_csv = params$guide_csv
has_guides = nchar(guide_csv) > 0 && file.exists(guide_csv)
if (has_guides) {
    guides = fread(guide_csv)
    guides[, barcode_clean := gsub("_", "", barcode)]
    setorder(guides, barcode_clean, -counts)
    guides_top = guides[, .SD[1L], by = barcode_clean]
    guides_top[, guide_gene := sub("_sgRNA.*", "", sgRNA)]

    for (pipe in names(seurat_objects)) {
        cells = colnames(seurat_objects[[pipe]])
        seurat_objects[[pipe]]$guide_assigned = ifelse(
            cells %in% guides_top$barcode_clean, "assigned", "unassigned")
    }
}
```

```{r umap-assign-celltype, eval=!is_hela}
ct_map = setNames(marker_meta$cell_type, marker_meta$marker)
label_map = setNames(
    paste0(marker_meta$marker, " (", marker_meta$cell_type, ")"),
    marker_meta$marker)

for (pipe in names(seurat_objects)) {
    so = seurat_objects[[pipe]]
    present = intersect(markers, rownames(so))
    if (length(present) == 0) {
        so$marker_celltype = "none"
    } else {
        expr_mat = GetAssayData(so, layer = "data")[present, , drop = FALSE]
        top_idx = apply(expr_mat, 2, which.max)
        top_val = apply(expr_mat, 2, max)
        top_marker = present[top_idx]
        assigned = unname(label_map[top_marker])
        assigned[top_val == 0] = "none"
        names(assigned) = colnames(expr_mat)
        so$marker_celltype = assigned
    }
    seurat_objects[[pipe]] = so
}

ct_levels = unname(c(label_map[sort(names(label_map))], "none"))
ct_colours = setNames(
    c(rep(cbbPalette, length.out = length(ct_levels) - 1), "grey85"),
    ct_levels)
```

```{r umap-clusters, results='asis', fig.width=ng_fig_size(length(seurat_objects), 1, panel_mm = 55)$w, fig.height=ng_fig_size(length(seurat_objects), 1, panel_mm = 55)$h}
cat("### Clusters\n\n")
cat("Each dot is a cell; colour encodes Louvain cluster. Nearby cells are",
    "transcriptionally similar. The number and separation of clusters shows",
    "how many cell types each pipeline detects.\n\n")
cluster_plots = lapply(names(seurat_objects), function(pipe) {
    so = seurat_objects[[pipe]]
    n_clusters = length(unique(so$cluster_prefixed))
    cluster_colours = rep(cbbPalette, ceiling(n_clusters / length(cbbPalette)))
    DimPlot(so, reduction = "umap", group.by = "cluster_prefixed",
            label = TRUE, label.size = 2.5,
            cols = cluster_colours[seq_len(n_clusters)]) +
        theme_ng() +
        labs(title = pipe, colour = "cluster")
})
print(wrap_plots(cluster_plots, nrow = 1))
```

```{r umap-pca, results='asis', fig.width=ng_fig_size(length(seurat_objects), 1, panel_mm = 55)$w, fig.height=ng_fig_size(length(seurat_objects), 1, panel_mm = 55)$h}
cat("### PCA\n\n")
cat("PC1 vs PC2 coloured by cluster. Compare with the UMAP to check whether",
    "the top PCs separate the same groups.\n\n")
pca_plots = lapply(names(seurat_objects), function(pipe) {
    so = seurat_objects[[pipe]]
    n_clusters = length(unique(so$cluster_prefixed))
    cluster_colours = rep(cbbPalette, ceiling(n_clusters / length(cbbPalette)))
    DimPlot(so, reduction = "pca", group.by = "cluster_prefixed",
            label = TRUE, label.size = 2.5,
            cols = cluster_colours[seq_len(n_clusters)]) +
        theme_ng() +
        labs(title = pipe, colour = "cluster")
})
print(wrap_plots(pca_plots, nrow = 1))
```

```{r umap-cellcycle, results='asis', fig.width=ng_fig_size(length(seurat_objects), 1, panel_mm = 55)$w, fig.height=ng_fig_size(length(seurat_objects), 1, panel_mm = 55)$h}
any_phase = any(vapply(seurat_objects,
                       function(so) any(!is.na(so$Phase)), logical(1)))
if (any_phase) {
    cat("### Cell cycle phase\n\n")
    cat("Each cell is assigned G1, S, or G2M by Seurat::CellCycleScoring using",
        "the Tirosh 2016 gene lists (human symbols, Title-cased for mouse).",
        "The same UMAP is shown as above, coloured by cell cycle phase.\n\n")
    phase_cols = c(G1 = "#999999", S = "#009E73", G2M = "#D55E00")
    phase_plots = lapply(names(seurat_objects), function(pipe) {
        so = seurat_objects[[pipe]]
        ph = factor(as.character(so$Phase), levels = c("G1", "S", "G2M"))
        DimPlot(so, reduction = "umap", group.by = "Phase",
                pt.size = 0.3, cols = phase_cols,
                shuffle = TRUE, seed = 1) +
            theme_ng() +
            labs(title = pipe, colour = "Cell cycle phase")
    })
    print(wrap_plots(phase_plots, nrow = 1) + plot_layout(guides = "collect"))
} else {
    cat("Cell cycle phase not assigned (no Tirosh gene overlap).\n")
}
```

```{r umap-celltype, eval=!is_hela, results='asis', fig.width=ng_fig_size(length(seurat_objects), 1, panel_mm = 55)$w, fig.height=ng_fig_size(length(seurat_objects), 1, panel_mm = 55)$h}
cat("### Cell type (marker-based)\n\n")
cat("For each cell, the marker with the highest log-normalised expression is",
    "taken and the cell is labelled with that marker's cell type. Cells with",
    "zero counts for all markers are labelled \"none\". Labels carry both the",
    "marker and the cell type so colours are directly interpretable.\n\n")
ct_plots = lapply(names(seurat_objects), function(pipe) {
    so = seurat_objects[[pipe]]
    so$marker_celltype = factor(so$marker_celltype, levels = ct_levels)
    DimPlot(so, reduction = "umap", group.by = "marker_celltype",
            pt.size = 0.3, cols = ct_colours,
            shuffle = TRUE, seed = 1) +
        theme_ng() +
        labs(title = pipe, colour = "cell type")
})
print(wrap_plots(ct_plots, nrow = 1) + plot_layout(guides = "collect"))
```

```{r umap-guide, results='asis', fig.width=ng_fig_size(length(seurat_objects), 1, panel_mm = 55)$w, fig.height=ng_fig_size(length(seurat_objects), 1, panel_mm = 55)$h}
if (has_guides) {
    cat("### Guide assignment\n\n")
    cat("Each cell is coloured by whether its barcode matched a guide assignment.",
        "Unassigned cells (red) are plotted on top. They either had no guide",
        "detected or their barcode did not match the author-provided table.\n\n")
    guide_plots = lapply(names(seurat_objects), function(pipe) {
        so = seurat_objects[[pipe]]
        so$guide_assigned = factor(so$guide_assigned,
                                   levels = c("assigned", "unassigned"))
        DimPlot(so, reduction = "umap", group.by = "guide_assigned",
                pt.size = 0.3, order = "unassigned",
                cols = c(assigned = "#4477AA", unassigned = "#CC3311")) +
            theme_ng() +
            labs(title = pipe, colour = NULL)
    })
    print(wrap_plots(guide_plots, nrow = 1) + plot_layout(guides = "collect"))
}
```

### Marker dotplot {.tabset}

Per-pipeline dotplot of marker expression across Louvain clusters. Dot size is
the fraction of cells in the cluster with raw count above zero; colour is the
scaled mean log-normalised expression (Seurat default). Clusters use the
per-pipeline prefix (a_N, s_N, k_N).

```{r umap-marker-dotplot, eval=!is_hela, results='asis', fig.width=ng_fig_size(1, 1, panel_mm = 80)$w, fig.height=ng_fig_size(1, 1, panel_mm = 80)$h}
for (pipe in names(seurat_objects)) {
    cat("####", pipe, "\n\n")
    so = seurat_objects[[pipe]]
    present = intersect(markers, rownames(so))
    if (length(present) == 0) {
        cat("No panel markers detected.\n\n")
        next
    }
    marker_labels = setNames(paste0(present, " (", ct_map[present], ")"), present)
    Idents(so) = factor(so$cluster_prefixed,
                        levels = sort(unique(so$cluster_prefixed)))
    p = DotPlot(so, features = present) +
        scale_x_discrete(labels = marker_labels) +
        scale_colour_viridis_c(option = "magma", direction = 1) +
        theme_ng() +
        theme(aspect.ratio = NULL,
              axis.text.x = element_text(angle = 30, hjust = 1)) +
        labs(x = NULL, y = NULL, title = pipe)
    print(p)
    cat("\n\n")
}
```

### Marker expression {.tabset}

```{r umap-markers, eval=!is_hela, results='asis', fig.width=ng_fig_size(length(seurat_objects), 1, panel_mm = 55)$w, fig.height=ng_fig_size(length(seurat_objects), 1, panel_mm = 55)$h}
for (i in seq_len(nrow(marker_meta))) {
    marker = marker_meta$marker[i]
    cell_type = marker_meta$cell_type[i]
    cat("####", marker, "\n\n")
    cat("Log-normalised expression of", marker, paste0("(", cell_type, ")"),
        "on the UMAP of each pipeline. Brighter cells carry higher expression;",
        "grey cells have zero counts.\n\n")
    feat_plots = lapply(names(seurat_objects), function(pipe) {
        so = seurat_objects[[pipe]]
        if (marker %in% rownames(so)) {
            FeaturePlot(so, reduction = "umap", features = marker,
                        order = TRUE, pt.size = 0.4) +
                scale_colour_viridis_c(option = "magma", direction = 1,
                                       na.value = "grey90") +
                theme_ng() +
                labs(title = pipe, subtitle = paste0(marker, " (", cell_type, ")"),
                     colour = "log norm")
        } else {
            ggplot() + theme_ng() +
                labs(title = pipe,
                     subtitle = paste(marker, "not detected"))
        }
    })
    print(wrap_plots(feat_plots, nrow = 1))
    cat("\n\n")
}
```

## Guide RNA assignments

This dataset is from a pooled CRISPR screen. The authors provide per-cell guide
assignments from their own processing of a separate amplicon library. We do not
quantify guides here; we use that table only as cell metadata. Barcodes in the
table use CB1_CB2_CB3 format (three 9-bp segments separated by underscores); we
remove underscores to match the 27-bp concatenated format this pipeline produces.
For cells with more than one detected guide we take the one with the highest UMI
count. The join rate is the fraction of cells from each aligner that match a
guide assignment; it shows how well each aligner's barcodes agree with the
original study.

```{r guide-assignments, results='asis', fig.width=ng_fig_size(length(seurat_objects), 1)$w, fig.height=ng_fig_size(length(seurat_objects), 1)$h}
if (has_guides) {
    join_stats = rbindlist(lapply(names(seurat_objects), function(pipe) {
        cells = colnames(seurat_objects[[pipe]])
        n_matched = sum(cells %in% guides_top$barcode_clean)
        data.table(
            pipeline = pipe,
            n_cells = length(cells),
            n_matched = n_matched,
            join_rate_pct = round(100 * n_matched / length(cells), 3)
        )
    }))

    saveRDS(join_stats, file.path(wd, paste0(samp, "_biology_guide_join.rds")))
    print(knitr::kable(join_stats,
        caption = paste("Fraction of pipeline cells matched to an author-assigned guide.",
                        "A higher join rate indicates better barcode concordance",
                        "with the original study's processing.")))

    print(
        ggplot(join_stats, aes(x = pipeline, y = join_rate_pct, fill = pipeline)) +
            geom_col() +
            geom_text(aes(label = paste0(join_rate_pct, "%")), vjust = -0.4, size = 3) +
            scale_fill_manual(values = pipeline_colours) +
            scale_y_continuous(limits = c(0, 110),
                               expand = expansion(mult = c(0, 0))) +
            theme_ng() +
            theme(aspect.ratio = NULL, legend.position = "none") +
            labs(x = NULL, y = "cells with guide assignment (%)",
                 title = "Guide barcode join rate per pipeline")
    )
} else {
    cat("No guide assignment file provided for this sample.\n")
}
```

```{r guide-top20, results='asis', fig.width=ng_fig_size(2, 1, panel_mm = 60)$w, fig.height=ng_fig_size(1, 1, panel_mm = 120)$h}
if (has_guides) {
    top_guides = guides_top[, .N, by = guide_gene][order(-N)][seq_len(min(20L, .N))]

    print(
        ggplot(top_guides, aes(x = reorder(guide_gene, N), y = N)) +
            geom_col(fill = cbbPalette[1]) +
            coord_flip() +
            theme_ng() +
            theme(aspect.ratio = NULL) +
            labs(x = NULL, y = "cells assigned",
                 title = "Top 20 guide targets (all barcodes)")
    )
}
```

## Marker expression per pipeline

A cell is counted as expressing a marker if its raw count exceeds zero. Mean
expression is computed only over expressing cells using log-normalised values.

```{r marker-expression-table, eval=!is_hela}
marker_expr = rbindlist(lapply(names(seurat_objects), function(pipe) {
    so = seurat_objects[[pipe]]
    n_total = ncol(so)
    rbindlist(lapply(seq_len(nrow(marker_meta)), function(i) {
        marker = marker_meta$marker[i]
        cell_type = marker_meta$cell_type[i]
        if (!(marker %in% rownames(so))) {
            return(data.table(
                pipeline = pipe,
                marker = marker,
                cell_type = cell_type,
                n_expressing = NA_integer_,
                pct_expressing = NA_real_,
                mean_lognorm_expr = NA_real_,
                n_total_cells = n_total
            ))
        }
        raw = GetAssayData(so, layer = "counts")[marker, ]
        norm = GetAssayData(so, layer = "data")[marker, ]
        n_pos = sum(raw > 0)
        data.table(
            pipeline = pipe,
            marker = marker,
            cell_type = cell_type,
            n_expressing = n_pos,
            pct_expressing = round(100 * n_pos / n_total, 2),
            mean_lognorm_expr = round(if (n_pos > 0) mean(norm[raw > 0]) else NA_real_, 3),
            n_total_cells = n_total
        )
    }))
}))

saveRDS(marker_expr, file.path(wd, paste0(samp, "_biology_marker_expr.rds")))
knitr::kable(
    marker_expr[order(marker, pipeline)],
    caption = paste(
        "Cells positive for each marker (raw count > 0) per pipeline.",
        "Mean log-normalised expression is computed over expressing cells only."
    )
)
```

```{r marker-expression-save}
cb_umi_dt = rbindlist(lapply(names(seurat_objects), function(pipe) {
    data.table(pipeline = pipe,
               barcode = colnames(seurat_objects[[pipe]]),
               total_umi = seurat_objects[[pipe]]$nCount_RNA)
}))
saveRDS(cb_umi_dt, file.path(wd, paste0(samp, "_biology_cb_umi.rds")))

cluster_dt = rbindlist(lapply(names(seurat_objects), function(pipe) {
    so = seurat_objects[[pipe]]
    ct = if (!is.null(so$marker_celltype)) as.character(so$marker_celltype)
         else rep(NA_character_, ncol(so))
    data.table(pipeline = pipe,
               barcode = colnames(so),
               cluster = as.character(Idents(so)),
               cluster_prefixed = so$cluster_prefixed,
               celltype = ct,
               Phase = as.character(so$Phase),
               S.Score = so$S.Score,
               G2M.Score = so$G2M.Score)
}))
saveRDS(cluster_dt, file.path(wd, paste0(samp, "_biology_clusters.rds")))
fwrite(cluster_dt, file.path(wd, paste0(samp, "_biology_clusters.csv")))

marker_sets = lapply(names(seurat_objects), function(pipe) {
    so = seurat_objects[[pipe]]
    present = intersect(markers, rownames(so))
    if (length(present) == 0) return(list(pipeline = pipe, sets = list()))
    counts_mat = GetAssayData(so, layer = "counts")[present, , drop = FALSE]
    cell_ids = paste0(pipe, "|", colnames(counts_mat))
    sets = lapply(seq_len(nrow(counts_mat)), function(i) {
        cell_ids[counts_mat[i, ] > 0]
    })
    names(sets) = paste0(present, " (", ct_map[present], ")")
    list(pipeline = pipe, sets = sets, cell_ids = cell_ids,
         counts_mat = counts_mat, present = present)
})
names(marker_sets) = names(seurat_objects)
saveRDS(marker_sets, file.path(wd, paste0(samp, "_biology_marker_sets.rds")))

n_pipelines = length(seurat_objects)
rm(seurat_objects)
gc()
```

## Cross-pipeline comparison

### Pseudobulk count deviation

For each pipeline, gene counts are summed across all cells to produce a
pseudobulk vector. Pairwise mean absolute relative difference (MARD) between
pipelines is shown as a heatmap. MARD = mean(|x - y| / ((x + y) / 2)) over
genes where at least one pipeline has non-zero counts. Lower values mean closer
agreement.

```{r pseudobulk-mard, fig.width=ng_fig_size(max(n_pipelines, 1), max(n_pipelines, 1))$w, fig.height=ng_fig_size(max(n_pipelines, 1), max(n_pipelines, 1))$h}
available_aligners = intersect(aligners, unique(cb_umi_dt$pipeline))

if (length(available_aligners) < 2) {
    cat("Pseudobulk MARD skipped: need at least 2 pipelines, have",
        length(available_aligners), "\n")
} else {
    sces_pb = lapply(available_aligners, function(aligner) {
        fn = file.path(wd, aligner, samp, paste0(samp, "_", aligner, "_sce.rds"))
        if (!file.exists(fn)) return(NULL)
        sce = readRDS(fn)
        sce = sce[rowSums(counts(sce)) > 0, colSums(counts(sce)) > 0]
        rd = rowData(sce)
        sym = rownames(sce)
        for (col in c("name", "SYMBOL", "gene_name")) {
            if (col %in% colnames(rd)) { sym = as.character(rd[[col]]); break }
        }
        rownames(sce) = make.unique(sym)
        rowSums(counts(sce))
    })
    names(sces_pb) = available_aligners
    sces_pb = Filter(Negate(is.null), sces_pb)

    shared_genes = Reduce(intersect, lapply(sces_pb, names))
    cat(sprintf("pseudobulk MARD: %d shared gene symbols across %d pipelines\n",
                length(shared_genes), length(sces_pb)))

    if (length(shared_genes) == 0 || length(sces_pb) < 2) {
        cat("Pseudobulk MARD skipped: no shared gene symbols across pipelines.",
            "Ensure rowData carries gene_name symbols for each aligner.\n")
    } else {
        pb_mat = do.call(cbind, lapply(sces_pb, function(v) v[shared_genes]))

        mard_mat = outer(
            colnames(pb_mat), colnames(pb_mat),
            Vectorize(function(a, b) {
                x = pb_mat[, a]
                y = pb_mat[, b]
                denom = (x + y) / 2
                mean(abs(x - y)[denom > 0] / denom[denom > 0], na.rm = TRUE)
            })
        )
        dimnames(mard_mat) = list(colnames(pb_mat), colnames(pb_mat))
        mard_long = as.data.frame(as.table(mard_mat))
        names(mard_long) = c("pipeline1", "pipeline2", "mard")

        saveRDS(list(mard = mard_mat, pb = pb_mat),
                file.path(wd, paste0(samp, "_biology_pseudobulk_mard.rds")))

        print(
            ggplot(mard_long, aes(pipeline1, pipeline2, fill = mard * 100)) +
                geom_tile(colour = "white") +
                geom_text(aes(label = sprintf("%.2f%%", mard * 100)), size = 3.2) +
                scale_fill_distiller(palette = "RdYlGn", direction = -1, limits = c(0, NA)) +
                theme_ng() +
                theme(aspect.ratio = NULL,
                      axis.text.x = element_text(angle = 30, hjust = 1)) +
                labs(title = "Pairwise pseudobulk MARD",
                     subtitle = "Lower is better",
                     x = NULL, y = NULL, fill = "MARD (%)")
        )
    }
}
```

### Pseudobulk correlation

Pairwise scatter plots of gene-level pseudobulk counts (log10 scale) for shared
genes. Each point is one gene; marker genes from the panel are highlighted.
Pearson correlation is computed on log10-transformed counts. A correlation near 1
means the two pipelines assign nearly identical total counts to each gene.

```{r pseudobulk-correlation, fig.width=ng_fig_size(3, 1, panel_mm = 70)$w, fig.height=ng_fig_size(3, 1, panel_mm = 70)$h}
if (exists("pb_mat") && ncol(pb_mat) >= 2) {
    pb_log = log10(pb_mat + 1)
    pb_dt = as.data.table(pb_log, keep.rownames = "gene")
    pb_dt[, is_marker := gene %in% markers]

    pairs_pb = combn(colnames(pb_mat), 2, simplify = FALSE)
    cor_plots = lapply(pairs_pb, function(pair) {
        d = pb_dt[, .(gene, x = get(pair[1]), y = get(pair[2]), is_marker)]
        r = cor(d$x, d$y)
        ggplot(d, aes(x = x, y = y)) +
            geom_point(data = d[is_marker == FALSE], alpha = 0.15, size = 0.3,
                       colour = "grey50") +
            geom_point(data = d[is_marker == TRUE], colour = cbbPalette[2],
                       size = 1.5) +
            geom_text_repel(data = d[is_marker == TRUE], aes(label = gene),
                            size = 2.6, colour = cbbPalette[2],
                            max.overlaps = Inf, min.segment.length = 0,
                            segment.size = 0.2, box.padding = 0.3) +
            geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                        colour = "grey30") +
            annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.3,
                     label = sprintf("R = %.4f", r), size = 3) +
            theme_ng() +
            theme(aspect.ratio = 1) +
            labs(x = paste0(pair[1], " log10(count + 1)"),
                 y = paste0(pair[2], " log10(count + 1)"),
                 title = paste(pair[1], "vs", pair[2]))
    })
    print(wrap_plots(cor_plots, nrow = 1))
} else {
    cat("Pseudobulk correlation skipped: need pb_mat with at least 2 pipelines.\n")
}
```

### Barcode overlap

How many cell barcodes are shared across pipelines. Bars unique to one aligner
show barcodes that only that aligner called. Intersections show barcodes called
by more than one aligner.

```{r barcode-overlap, fig.width=8, fig.height=5}
bc_lists = split(cb_umi_dt$barcode, cb_umi_dt$pipeline)
if (length(bc_lists) < 2) {
    cat("Barcode overlap skipped: need at least 2 pipelines, have",
        length(bc_lists), "\n")
} else {
    upset(fromList(bc_lists), order.by = "freq", text.scale = 1.4,
          point.size = 3, line.size = 0.8, mb.ratio = c(0.6, 0.4))
}
```

### Per-barcode UMI concordance

For each pair of pipelines, only barcodes detected by both are compared.
Mean UMI (x-axis) and difference in UMI (y-axis) follow the Bland-Altman
convention. Dashed lines show mean bias and 1.96 SD limits of agreement.

```{r bland-altman-umi, fig.width=ng_fig_size(3, 1, panel_mm = 70)$w, fig.height=ng_fig_size(3, 1, panel_mm = 70)$h}
pipes = unique(cb_umi_dt$pipeline)

if (length(pipes) < 2) {
    cat("Bland-Altman skipped: need at least 2 pipelines, have", length(pipes), "\n")
} else {
    pairs = combn(pipes, 2, simplify = FALSE)
    ba_plots = lapply(pairs, function(pair) {
        d1 = cb_umi_dt[pipeline == pair[1], .(barcode, umi1 = total_umi)]
        d2 = cb_umi_dt[pipeline == pair[2], .(barcode, umi2 = total_umi)]
        shared = merge(d1, d2, by = "barcode")
        shared[, mean_umi := (log10(umi1 + 1) + log10(umi2 + 1)) / 2]
        shared[, diff_umi := log10(umi1 + 1) - log10(umi2 + 1)]
        bias = mean(shared$diff_umi)
        loa  = sd(shared$diff_umi) * 1.96
        ggplot(shared, aes(x = mean_umi, y = diff_umi)) +
            geom_point(alpha = 0.1, size = 0.3) +
            geom_hline(yintercept = bias, colour = cbbPalette[2], linewidth = 0.6) +
            geom_hline(yintercept = c(bias - loa, bias + loa),
                       colour = cbbPalette[2], linewidth = 0.4, linetype = "dashed") +
            theme_ng() +
            theme(aspect.ratio = 1) +
            labs(x = "mean log10(UMI + 1)",
                 y = "diff log10(UMI + 1)",
                 title = paste(pair[1], "vs", pair[2]),
                 subtitle = sprintf("n = %d shared barcodes, bias = %.3f",
                                    nrow(shared), bias))
    })
    print(wrap_plots(ba_plots, nrow = 1))
}
```

### Cell recovery

Total cells called by each pipeline. Differences come from each aligner's
barcode calling and filtering, not from biology; these numbers are the denominator
for the percentage calculations below.

```{r cell-recovery, eval=!is_hela, fig.width=ng_fig_size(n_pipelines, 1)$w, fig.height=ng_fig_size(n_pipelines, 1)$h}
ncell_df = unique(marker_expr[, .(pipeline, n_total_cells)])

ggplot(ncell_df, aes(x = pipeline, y = n_total_cells, fill = pipeline)) +
    geom_col() +
    geom_text(aes(label = n_total_cells), vjust = -0.4, size = 3) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    scale_fill_manual(values = pipeline_colours) +
    theme_ng() +
    theme(aspect.ratio = NULL, legend.position = "none") +
    labs(x = NULL, y = "cells recovered")
```

### Marker counts

Per-pipeline table of cells positive for each marker (raw count above zero) and
mean log-normalised expression over those positive cells. Table above, bars
below.

```{r marker-count-table, eval=!is_hela, results='asis'}
print(knitr::kable(
    marker_expr[order(marker, pipeline)],
    caption = paste(
        "Cells positive for each marker (raw count > 0) per pipeline.",
        "Mean log-normalised expression is computed over expressing cells only."
    )
))
```

Number of cells with at least one UMI for each marker, split by pipeline. A
higher count does not always mean better detection; it can also come from more
cells being called or spurious alignment.

```{r marker-count-plot, eval=!is_hela, fig.width=ng_fig_size(max(length(markers), 1L), 1, panel_mm = 25)$w, fig.height=ng_fig_size(max(length(markers), 1L), 3, panel_mm = 25)$h}
expr_ok = marker_expr[!is.na(n_expressing)]
expr_ok[, marker_ct := paste0(marker, "\n", cell_type)]

ggplot(expr_ok, aes(x = marker_ct, y = n_expressing, fill = pipeline)) +
    geom_col(position = "dodge") +
    scale_fill_manual(values = pipeline_colours) +
    theme_ng() +
    theme(aspect.ratio = NULL) +
    labs(x = NULL, y = "cells expressing (count > 0)", fill = NULL)
```

Fraction of cells that are positive for each marker. Dividing by cell count
removes the effect of each pipeline calling different numbers of cells, so the
pipelines can be compared on the same scale.

```{r marker-pct-plot, eval=!is_hela, fig.width=ng_fig_size(max(length(markers), 1L), 1, panel_mm = 25)$w, fig.height=ng_fig_size(max(length(markers), 1L), 3, panel_mm = 25)$h}
ggplot(expr_ok, aes(x = marker_ct, y = pct_expressing, fill = pipeline)) +
    geom_col(position = "dodge") +
    scale_fill_manual(values = pipeline_colours) +
    theme_ng() +
    theme(aspect.ratio = NULL) +
    labs(x = NULL, y = "percent expressing (%)", fill = NULL)
```

### Marker co-expression (UpSet) {.tabset}

For each pipeline, cells are grouped by which panel markers they express (raw
count above zero). Bars show the number of cells in each intersection.

```{r marker-upset, eval=!is_hela, results='asis', fig.width=7, fig.height=5}
for (pipe in names(marker_sets)) {
    cat("####", pipe, "\n\n")
    ms = marker_sets[[pipe]]
    if (length(ms$sets) == 0) {
        cat("No panel markers detected.\n\n")
        next
    }
    sets_named = ms$sets
    names(sets_named) = ms$present
    non_empty = sets_named[vapply(sets_named, length, integer(1)) > 0]
    if (length(non_empty) < 2) {
        cat("Fewer than two markers with positive cells; skipping.\n\n")
        next
    }
    print(upset(fromList(non_empty), order.by = "freq",
                nsets = length(non_empty), nintersects = 30,
                text.scale = 1.2, point.size = 2.5, line.size = 0.6))
    cat("\n\n")
}
```

### Marker combinations per pipeline

For each cell we record the set of markers it co-expresses. The top 15
combinations (by overall cell count) are shown as grouped bars; all remaining
combinations are pooled into "other". Bars show the percentage of each
pipeline's cells in that combination, so pipelines with different totals are
directly comparable. Cells with zero counts across all markers are labelled
"none".

```{r marker-combo-bar, eval=!is_hela, fig.width=ng_fig_size(3, 1, panel_mm = 70)$w, fig.height=ng_fig_size(3, 1, panel_mm = 70)$h * 2}
combo_dt = rbindlist(lapply(names(marker_sets), function(pipe) {
    ms = marker_sets[[pipe]]
    if (length(ms$present) == 0 || is.null(ms$counts_mat)) {
        n_cells = if (is.null(ms$cell_ids)) 0L else length(ms$cell_ids)
        if (n_cells == 0) return(NULL)
        return(data.table(pipeline = pipe, combo = rep("none", n_cells)))
    }
    labels = paste0(ms$present, "(", ct_map[ms$present], ")")
    membership = ms$counts_mat > 0
    combo_per_cell = apply(membership, 2, function(col) {
        hits = labels[col]
        if (length(hits) == 0) "none" else paste(hits, collapse = "+")
    })
    data.table(pipeline = pipe, combo = combo_per_cell)
}), fill = TRUE)

pipe_totals = combo_dt[, .(n_total = .N), by = pipeline]
combo_counts = combo_dt[, .(n = .N), by = .(pipeline, combo)]

overall = combo_counts[, .(n_all = sum(n)), by = combo][order(-n_all)]
top_combos = overall$combo[seq_len(min(15L, nrow(overall)))]
combo_counts[, combo_capped := ifelse(combo %in% top_combos, combo, "other")]

plot_dt = combo_counts[, .(n = sum(n)), by = .(pipeline, combo_capped)]
plot_dt = merge(plot_dt, pipe_totals, by = "pipeline", all.x = TRUE)
plot_dt[, pct := 100 * n / n_total]

combo_order = c(top_combos[top_combos %in% unique(plot_dt$combo_capped)])
if ("other" %in% plot_dt$combo_capped) combo_order = c(combo_order, "other")
plot_dt[, combo_capped := factor(combo_capped, levels = combo_order)]

ggplot(plot_dt, aes(x = combo_capped, y = pct, fill = pipeline)) +
    geom_col(position = "dodge") +
    scale_fill_manual(values = pipeline_colours) +
    theme_ng() +
    theme(aspect.ratio = NULL,
          axis.text.x = element_text(angle = 35, hjust = 1)) +
    labs(x = NULL, y = "cells in combination (% of pipeline total)",
         fill = NULL, title = "Marker combinations per pipeline (top 15 + other)")
```

### Dotplot summary

Dot size: fraction of cells with at least one count for the marker. Colour: mean
log-normalised expression in those cells. A small bright dot means the marker is
rare but highly expressed; a large pale dot means it is widespread but lowly
expressed. Rows are pipelines; columns are markers.

```{r dotplot, eval=!is_hela, fig.width=ng_fig_size(max(length(markers), 1L), n_pipelines, panel_mm = 20)$w, fig.height=ng_fig_size(max(length(markers), 1L), n_pipelines, panel_mm = 20)$h}
expr_ok[, marker_label := paste0(marker, "\n", cell_type)]

ggplot(expr_ok,
       aes(x = marker_label, y = pipeline,
           size = pct_expressing,
           colour = mean_lognorm_expr)) +
    geom_point() +
    scale_colour_viridis_c(option = "magma", direction = 1, na.value = "grey90") +
    scale_size_continuous(range = c(0.5, 6)) +
    theme_ng() +
    theme(aspect.ratio = NULL) +
    labs(x = NULL, y = NULL,
         size = "% expressing",
         colour = "mean log norm")
```

### Cluster agreement (adjusted Rand index)

Each pipeline clusters cells independently (Louvain, resolution 0.5). For each
pair of pipelines, only barcodes detected by both are kept. Because cluster
labels are arbitrary (cluster 3 in starsolo may correspond to cluster 7 in
kallisto), the Hungarian algorithm is used to find the optimal one-to-one mapping
between cluster labels that maximises agreement. The adjusted Rand index (ARI) is
then computed on the original (unmapped) labels; ARI = 1 means perfect agreement,
ARI = 0 means no better than random. The confusion matrix after Hungarian
relabelling is shown as a heatmap for each pair.

```{r cluster-ari, fig.width=ng_fig_size(3, 1, panel_mm = 70)$w, fig.height=ng_fig_size(3, 1, panel_mm = 70)$h}
square_pad = function(ct) {
    nr = nrow(ct); nc = ncol(ct)
    if (nr == nc) return(ct)
    if (nr > nc) {
        extra = matrix(0L, nrow = nr, ncol = nr - nc,
                       dimnames = list(rownames(ct),
                                       paste0("__pad", seq_len(nr - nc))))
        cbind(ct, extra)
    } else {
        extra = matrix(0L, nrow = nc - nr, ncol = nc,
                       dimnames = list(paste0("__pad", seq_len(nc - nr)),
                                       colnames(ct)))
        rbind(ct, extra)
    }
}

if (!requireNamespace("clue", quietly = TRUE)) {
    cat("Package 'clue' not available; skipping cluster ARI.\n")
} else if (nrow(cluster_dt) == 0 || length(unique(cluster_dt$pipeline)) < 2) {
    cat("Cluster ARI skipped: need at least 2 pipelines.\n")
} else {
    pipes_cl = unique(cluster_dt$pipeline)
    pairs_cl = combn(pipes_cl, 2, simplify = FALSE)

    ari_results = rbindlist(lapply(pairs_cl, function(pair) {
        d1 = cluster_dt[pipeline == pair[1], .(barcode, cl1 = cluster_prefixed)]
        d2 = cluster_dt[pipeline == pair[2], .(barcode, cl2 = cluster_prefixed)]
        shared = merge(d1, d2, by = "barcode")
        if (nrow(shared) < 10) return(NULL)
        ari = mclust::adjustedRandIndex(shared$cl1, shared$cl2)
        data.table(pipeline1 = pair[1], pipeline2 = pair[2],
                   n_shared = nrow(shared), ari = round(ari, 4))
    }))

    saveRDS(ari_results, file.path(wd, paste0(samp, "_biology_cluster_ari.rds")))
    print(knitr::kable(ari_results,
        caption = "Adjusted Rand index for cluster labels on shared barcodes."))

    ari_plots = lapply(pairs_cl, function(pair) {
        d1 = cluster_dt[pipeline == pair[1], .(barcode, cl1 = cluster_prefixed)]
        d2 = cluster_dt[pipeline == pair[2], .(barcode, cl2 = cluster_prefixed)]
        shared = merge(d1, d2, by = "barcode")
        if (nrow(shared) < 10) return(NULL)

        ct = table(shared$cl1, shared$cl2)
        ct_sq = square_pad(ct)
        mapping = clue::solve_LSAP(max(ct_sq) - ct_sq)
        new_order_full = colnames(ct_sq)[mapping]
        new_order = new_order_full[new_order_full %in% colnames(ct)]

        ct_dt = as.data.table(as.table(ct))
        names(ct_dt) = c("cluster_a", "cluster_b", "n")
        ct_dt[, cluster_b := factor(cluster_b, levels = new_order)]

        ari_val = ari_results[pipeline1 == pair[1] & pipeline2 == pair[2], ari]
        ggplot(ct_dt, aes(x = cluster_a, y = cluster_b, fill = n)) +
            geom_tile(colour = "white") +
            geom_text(aes(label = ifelse(n > 0, n, "")), size = 2.2) +
            scale_fill_distiller(palette = "Blues", direction = 1) +
            theme_ng() +
            theme(aspect.ratio = 1,
                  axis.text.x = element_text(angle = 30, hjust = 1)) +
            labs(x = pair[1], y = paste0(pair[2], " (reordered)"),
                 title = paste(pair[1], "vs", pair[2]),
                 subtitle = sprintf("ARI = %.4f, n = %d shared cells",
                                    ari_val, nrow(shared)),
                 fill = "cells")
    })
    ari_plots = Filter(Negate(is.null), ari_plots)
    if (length(ari_plots) > 0) print(wrap_plots(ari_plots, nrow = 1))
}
```

### Cell-type agreement (adjusted Rand index)

Each cell is assigned to a marker-based cell type using the highest-expressing
panel marker (same rule as the UMAP panel above). Because the label vocabulary
is shared across pipelines (one label per marker, plus "none"), ARI is computed
directly on these labels without Hungarian relabelling. The confusion matrix
between the two pipelines is shown for each pair.

```{r celltype-ari, eval=!is_hela, fig.width=ng_fig_size(3, 1, panel_mm = 70)$w * 2, fig.height=ng_fig_size(3, 1, panel_mm = 70)$h * 2}
if (nrow(cluster_dt) == 0 || length(unique(cluster_dt$pipeline)) < 2) {
    cat("Cell-type ARI skipped: need at least 2 pipelines.\n")
} else {
    pipes_ct = unique(cluster_dt$pipeline)
    pairs_ct = combn(pipes_ct, 2, simplify = FALSE)

    ct_ari_results = rbindlist(lapply(pairs_ct, function(pair) {
        d1 = cluster_dt[pipeline == pair[1], .(barcode, ct1 = celltype)]
        d2 = cluster_dt[pipeline == pair[2], .(barcode, ct2 = celltype)]
        shared = merge(d1, d2, by = "barcode")
        if (nrow(shared) < 10) return(NULL)
        ari = mclust::adjustedRandIndex(shared$ct1, shared$ct2)
        data.table(pipeline1 = pair[1], pipeline2 = pair[2],
                   n_shared = nrow(shared), ari = round(ari, 4))
    }))

    saveRDS(ct_ari_results, file.path(wd, paste0(samp, "_biology_celltype_ari.rds")))
    print(knitr::kable(ct_ari_results,
        caption = "Adjusted Rand index for marker-based cell type labels on shared barcodes."))

    ct_ari_plots = lapply(pairs_ct, function(pair) {
        d1 = cluster_dt[pipeline == pair[1], .(barcode, ct1 = celltype)]
        d2 = cluster_dt[pipeline == pair[2], .(barcode, ct2 = celltype)]
        shared = merge(d1, d2, by = "barcode")
        if (nrow(shared) < 10) return(NULL)

        ct = table(shared$ct1, shared$ct2)
        ct_dt = as.data.table(as.table(ct))
        names(ct_dt) = c("celltype_a", "celltype_b", "n")

        ari_val = ct_ari_results[pipeline1 == pair[1] & pipeline2 == pair[2], ari]
        ggplot(ct_dt, aes(x = celltype_a, y = celltype_b, fill = n)) +
            geom_tile(colour = "white") +
            geom_text(aes(label = ifelse(n > 0, n, "")), size = 2.2) +
            scale_fill_distiller(palette = "Blues", direction = 1) +
            theme_ng() +
            theme(aspect.ratio = 1,
                  axis.text.x = element_text(angle = 30, hjust = 1)) +
            labs(x = pair[1], y = pair[2],
                 title = paste(pair[1], "vs", pair[2]),
                 subtitle = sprintf("ARI = %.4f, n = %d shared cells",
                                    ari_val, nrow(shared)),
                 fill = "cells")
    })
    ct_ari_plots = Filter(Negate(is.null), ct_ari_plots)
    if (length(ct_ari_plots) > 0) print(wrap_plots(ct_ari_plots, nrow = 1))
}
```

## Computational performance

Wall-clock time, peak memory (max RSS), and mean CPU load for each pipeline step.
Benchmark files are produced by Snakemake and stored in the working directory.
Only aligner-specific steps are shown; shared preprocessing steps (indexing,
barcode standardisation) are excluded because they run once regardless of aligner.

```{r load-benchmarks-bio}
bm_dir = file.path(wd, "benchmarks")
bm_files = list.files(bm_dir, pattern = "\\.txt$", full.names = TRUE)

if (length(bm_files) == 0) {
    cat("No benchmark files found in", bm_dir, "\n")
} else {
    bm_raw = rbindlist(Filter(Negate(is.null), lapply(bm_files, function(fn) {
        tryCatch({
            dt = fread(fn, sep = "\t")
            dt[, file := basename(fn)]
            dt
        }, error = function(e) NULL)
    })), fill = TRUE)

    bm_raw[, step := gsub("\\.txt$", "", file)]

    classify_pipe = function(x) {
        ifelse(grepl("starsolo|^star_", x), "starsolo",
        ifelse(grepl("kallisto|bustools", x), "kallisto",
        ifelse(grepl("alevin", x), "alevin",
        ifelse(grepl("sbg", x), "sbg", "shared"))))
    }
    bm_raw[, pipeline := classify_pipe(step)]
    bm_pipe = bm_raw[pipeline %in% aligners]
    bm_pipe[, minutes := s / 60]
    bm_pipe[, max_rss_gb := max_rss / 1024]
}
```

```{r perf-time, fig.width=ng_fig_size(max(length(aligners), 1), 1, panel_mm = 60)$w * 0.7, fig.height=ng_fig_size(max(length(aligners), 1), 1, panel_mm = 60)$h}
if (exists("bm_pipe") && nrow(bm_pipe) > 0) {
    pipe_time = bm_pipe[, .(total_min = sum(minutes)), by = pipeline]

    print(
        ggplot(pipe_time, aes(x = pipeline, y = total_min, fill = pipeline)) +
            geom_col(show.legend = FALSE) +
            geom_text(aes(label = sprintf("%.1f", total_min)), vjust = -0.4, size = 3) +
            scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
            scale_fill_manual(values = pipeline_colours) +
            theme_ng() +
            theme(aspect.ratio = NULL) +
            labs(x = NULL, y = "wall-clock time (min)",
                 title = "Total runtime per pipeline")
    )
}
```

```{r perf-memory, fig.width=ng_fig_size(max(length(aligners), 1), 1, panel_mm = 60)$w * 0.7, fig.height=ng_fig_size(max(length(aligners), 1), 1, panel_mm = 60)$h}
if (exists("bm_pipe") && nrow(bm_pipe) > 0) {
    pipe_mem = bm_pipe[, .(peak_rss_gb = max(max_rss_gb, na.rm = TRUE)), by = pipeline]

    print(
        ggplot(pipe_mem, aes(x = pipeline, y = peak_rss_gb, fill = pipeline)) +
            geom_col(show.legend = FALSE) +
            geom_text(aes(label = sprintf("%.1f", peak_rss_gb)), vjust = -0.4, size = 3) +
            scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
            scale_fill_manual(values = pipeline_colours) +
            theme_ng() +
            theme(aspect.ratio = NULL) +
            labs(x = NULL, y = "peak RSS (GB)",
                 title = "Peak memory per pipeline")
    )
}
```

```{r perf-step-detail, results='asis', fig.width=7, fig.height=ng_fig_size(1, max(nrow(bm_pipe), 1), panel_mm = 12)$h * 0.5}
if (exists("bm_pipe") && nrow(bm_pipe) > 0) {
    print(
        ggplot(bm_pipe, aes(x = reorder(step, minutes), y = minutes, fill = pipeline)) +
            geom_col(width = 0.7) +
            coord_flip() +
            scale_fill_manual(values = pipeline_colours) +
            theme_ng() +
            theme(aspect.ratio = NULL) +
            labs(x = NULL, y = "wall-clock time (min)", fill = NULL,
                 title = "Runtime per step")
    )

    print(knitr::kable(
        bm_pipe[order(pipeline, -s),
                .(step, pipeline,
                  minutes = round(minutes, 2),
                  peak_rss_gb = round(max_rss_gb, 2),
                  mean_cpu_load = round(mean_load, 1))],
        caption = "Per-step benchmark summary (aligner-specific steps only)."
    ))
}
```

## Session info

```{r session-info}
sessionInfo()
```
