---
title: "rhapsodist: biology report"
author: "Izaskun Mallona"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    self_contained: true
params:
  working_dir: "/tmp/rhapsodist_simulate"
  sample: "simulated"
  aligners: "starsolo,kallisto,alevin"
  has_sbg: false
  markers_file: ""
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
    error = TRUE,
    echo = TRUE,
    cache = FALSE,
    dev = "png",
    dev.args = list(type = "cairo-png"),
    warning = FALSE,
    message = FALSE
)

suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(Seurat)
    library(scuttle)
    library(ggplot2)
    library(patchwork)
    library(data.table)
    library(Matrix)
    library(viridis)
})

source(file.path(dirname(knitr::current_input()), "..", "ggtheme.R"))

wd = params$working_dir
samp = params$sample
aligners = strsplit(params$aligners, ",")[[1]]
has_sbg = isTRUE(params$has_sbg) || identical(params$has_sbg, "true")

marker_meta = read.table(params$markers_file, header = TRUE, sep = "\t",
                         stringsAsFactors = FALSE)
markers = marker_meta$marker

if (has_sbg) aligners = c(aligners, "sbg")
```

## Data loading

Each pipeline produces a SingleCellExperiment saved as an RDS file. Gene symbols
are stored in rowData under different column names depending on the aligner
(starsolo: `name`, kallisto: `SYMBOL`, alevin/tximeta: Ensembl IDs as rownames
with no symbol column). We rename SCE rownames to symbols before building Seurat
objects so that marker lookups work uniformly across pipelines.

```{r load-sces}
load_sce = function(aligner, wd, samp) {
    fn = file.path(wd, aligner, samp, paste0(samp, "_", aligner, "_sce.rds"))
    if (!file.exists(fn)) {
        warning("SCE not found: ", fn)
        return(NULL)
    }
    readRDS(fn)
}

sces = setNames(lapply(aligners, load_sce, wd = wd, samp = samp), aligners)
sces = Filter(Negate(is.null), sces)

rm_zeros = function(sce) {
    sce = sce[rowSums(counts(sce)) > 0, ]
    sce[, colSums(counts(sce)) > 0]
}
sces = lapply(sces, rm_zeros)

symbol_rownames = function(sce) {
    rd = rowData(sce)
    sym = rownames(sce)
    for (col in c("name", "SYMBOL", "gene_name")) {
        if (col %in% colnames(rd)) {
            sym = as.character(rd[[col]])
            break
        }
    }
    rownames(sce) = make.unique(sym)
    sce
}

sces = lapply(sces, symbol_rownames)

cat("Pipelines loaded:", paste(names(sces), collapse = ", "), "\n")
sapply(sces, dim) |> `rownames<-`(c("n_genes", "n_cells"))
```

## QC summary

Cell and library quality metrics for each pipeline. Differences in cell count
reflect differences in barcode calling strategies. Median UMI and genes detected
reflect the sensitivity of each aligner.

```{r qc-table}
qc_df = rbindlist(lapply(names(sces), function(m) {
    sce = addPerCellQCMetrics(sces[[m]],
        subsets = list(mito = grep("^mt-|^MT-", rownames(sces[[m]]), value = TRUE)))
    data.table(
        pipeline = m,
        n_cells = ncol(sce),
        n_genes_total = nrow(sce),
        median_umi = round(median(sce$sum)),
        median_genes = round(median(sce$detected)),
        median_pct_mito = round(median(sce$subsets_mito_percent), 2)
    )
}))

saveRDS(qc_df, file.path(wd, paste0(samp, "_biology_qc.rds")))
knitr::kable(qc_df, caption = "Per-pipeline QC summary.")
```

Cells per pipeline, median UMI per cell, and median genes detected per cell.
Bars allow quick visual comparison; exact values are in the table above.

```{r qc-barplots, fig.width=ng_fig_size(3, 1)$w, fig.height=ng_fig_size(3, 1)$h}
qc_long = melt(
    qc_df[, .(pipeline, n_cells, median_umi, median_genes)],
    id.vars = "pipeline", variable.name = "metric", value.name = "value"
)

ggplot(qc_long, aes(x = pipeline, y = value, fill = pipeline)) +
    geom_col() +
    geom_text(aes(label = value), vjust = -0.4, size = 2.5) +
    facet_wrap(~metric, scales = "free_y") +
    scale_fill_manual(values = cbbPalette) +
    theme_ng() +
    theme(aspect.ratio = NULL, legend.position = "none") +
    labs(x = NULL, y = NULL)
```

Per-cell UMI and gene count distributions across pipelines. Rightward shifts
indicate higher library complexity. Multimodal distributions can reflect doublets
or heterogeneous cell populations captured at different efficiencies.

```{r umi-gene-density, fig.width=ng_fig_size(2, 1)$w, fig.height=ng_fig_size(2, 1)$h}
umi_df = rbindlist(lapply(names(sces), function(m) {
    data.table(pipeline = m, umi = colSums(counts(sces[[m]])))
}))

gene_df = rbindlist(lapply(names(sces), function(m) {
    data.table(pipeline = m, genes = colSums(counts(sces[[m]]) > 0))
}))

p_umi = ggplot(umi_df, aes(x = log10(umi + 1), colour = pipeline)) +
    geom_density(linewidth = 0.6) +
    scale_colour_manual(values = cbbPalette) +
    theme_ng() +
    theme(aspect.ratio = NULL) +
    labs(x = "log10(UMI + 1)", y = "density", colour = NULL)

p_gene = ggplot(gene_df, aes(x = log10(genes + 1), colour = pipeline)) +
    geom_density(linewidth = 0.6) +
    scale_colour_manual(values = cbbPalette) +
    theme_ng() +
    theme(aspect.ratio = NULL) +
    labs(x = "log10(genes + 1)", y = "density", colour = NULL)

p_umi + p_gene + plot_layout(guides = "collect")
```

## Per-pipeline clustering and marker expression {.tabset}

For each pipeline: 2000 highly variable genes selected by VST, 30 PCs, Louvain
clustering at resolution 0.5, and UMAP embedding using the top 20 PCs. Feature
plots show log-normalised expression; cells with zero counts for that gene are
shown in grey.

```{r build-seurat-objects}
make_seurat = function(sce) {
    so = CreateSeuratObject(counts = counts(sce), project = mainExpName(sce),
                            min.cells = 0, min.features = 0)
    so = NormalizeData(so, normalization.method = "LogNormalize",
                       scale.factor = 1e4, verbose = FALSE)
    so = FindVariableFeatures(so, selection.method = "vst", nfeatures = 2000,
                              verbose = FALSE)
    so = ScaleData(so, features = rownames(so), verbose = FALSE)
    so = RunPCA(so, features = VariableFeatures(so), npcs = 30, verbose = FALSE)
    so = FindNeighbors(so, dims = 1:20, verbose = FALSE)
    so = FindClusters(so, resolution = 0.5, verbose = FALSE)
    so = RunUMAP(so, dims = 1:20, verbose = FALSE)
    so
}

seurat_objects = lapply(sces, make_seurat)

# save each pipeline separately so individual objects can be loaded for debugging
for (pipe in names(seurat_objects)) {
    saveRDS(seurat_objects[[pipe]],
            file.path(wd, paste0(samp, "_biology_", pipe, "_seurat.rds")))
}

rm(sces)
gc()
```

```{r per-pipeline-tabs, results='asis', fig.width=ng_fig_size(1, 1)$w, fig.height=ng_fig_size(1, 1)$h}
for (pipe in names(seurat_objects)) {
    so = seurat_objects[[pipe]]
    n_clusters = length(unique(Idents(so)))
    cluster_colours = rep(cbbPalette, ceiling(n_clusters / length(cbbPalette)))

    cat("### ", pipe, " {.tabset}\n\n")

    cat("#### UMAP\n\n")
    cat("Each dot is a cell; colour encodes Louvain cluster. Proximity reflects",
        "transcriptomic similarity. The number and separation of clusters reflects",
        "the cell-type diversity recovered by this pipeline.\n\n")
    print(
        DimPlot(so, reduction = "umap", label = TRUE, label.size = 2.5,
                cols = cluster_colours[seq_len(n_clusters)]) +
        theme_ng() +
        labs(title = pipe, colour = "cluster")
    )
    cat("\n\n")

    cat("#### PCA\n\n")
    cat("PC1 vs PC2 coloured by cluster. Compare with the UMAP to check whether",
        "the top principal components capture the dominant sources of variation",
        "in this pipeline's gene set.\n\n")
    print(
        DimPlot(so, reduction = "pca", label = TRUE, label.size = 2.5,
                cols = cluster_colours[seq_len(n_clusters)]) +
        theme_ng() +
        labs(title = pipe, colour = "cluster")
    )
    cat("\n\n")

    for (i in seq_len(nrow(marker_meta))) {
        marker = marker_meta$marker[i]
        cell_type = marker_meta$cell_type[i]
        cat("####", marker, "\n\n")
        if (marker %in% rownames(so)) {
            cat("Log-normalised expression of", marker,
                paste0("(", cell_type, ")"),
                "on the", pipe, "UMAP.",
                "Brighter cells carry higher expression;",
                "grey cells have zero counts for this gene.\n\n")
            print(
                FeaturePlot(so, reduction = "umap", features = marker,
                            order = TRUE, pt.size = 0.4) +
                scale_colour_viridis_c(option = "magma", direction = 1,
                                       na.value = "grey90") +
                theme_ng() +
                labs(title = paste0(marker, " (", cell_type, ")"),
                     subtitle = pipe,
                     colour = "log norm")
            )
        } else {
            cat(marker, "not detected in the", pipe, "feature set.\n")
        }
        cat("\n\n")
    }
}
```

## Marker expression per pipeline

A cell is counted as expressing a marker if its raw count exceeds zero. Mean
expression is computed only over expressing cells using log-normalised values.

```{r marker-expression-table}
marker_expr = rbindlist(lapply(names(seurat_objects), function(pipe) {
    so = seurat_objects[[pipe]]
    n_total = ncol(so)
    rbindlist(lapply(seq_len(nrow(marker_meta)), function(i) {
        marker = marker_meta$marker[i]
        cell_type = marker_meta$cell_type[i]
        if (!(marker %in% rownames(so))) {
            return(data.table(
                pipeline = pipe,
                marker = marker,
                cell_type = cell_type,
                n_expressing = NA_integer_,
                pct_expressing = NA_real_,
                mean_lognorm_expr = NA_real_,
                n_total_cells = n_total
            ))
        }
        raw = GetAssayData(so, layer = "counts")[marker, ]
        norm = GetAssayData(so, layer = "data")[marker, ]
        n_pos = sum(raw > 0)
        data.table(
            pipeline = pipe,
            marker = marker,
            cell_type = cell_type,
            n_expressing = n_pos,
            pct_expressing = round(100 * n_pos / n_total, 2),
            mean_lognorm_expr = round(if (n_pos > 0) mean(norm[raw > 0]) else NA_real_, 3),
            n_total_cells = n_total
        )
    }))
}))

saveRDS(marker_expr, file.path(wd, paste0(samp, "_biology_marker_expr.rds")))
knitr::kable(
    marker_expr[order(marker, pipeline)],
    caption = paste(
        "Cells positive for each marker (raw count > 0) per pipeline.",
        "Mean log-normalised expression is computed over expressing cells only."
    )
)

n_pipelines = length(seurat_objects)
rm(seurat_objects)
gc()
```

## Cross-pipeline comparison

### Cell recovery

Total cells recovered by each pipeline. Differences reflect barcode calling and
filtering strategies, not biology; they set the denominator for all downstream
percentage calculations.

```{r cell-recovery, fig.width=ng_fig_size(n_pipelines, 1)$w, fig.height=ng_fig_size(n_pipelines, 1)$h}
ncell_df = unique(marker_expr[, .(pipeline, n_total_cells)])

ggplot(ncell_df, aes(x = pipeline, y = n_total_cells, fill = pipeline)) +
    geom_col() +
    geom_text(aes(label = n_total_cells), vjust = -0.4, size = 3) +
    scale_fill_manual(values = cbbPalette) +
    theme_ng() +
    theme(aspect.ratio = NULL, legend.position = "none") +
    labs(x = NULL, y = "cells recovered")
```

### Marker counts

Number of cells with at least one UMI for each marker, split by pipeline. A
higher count does not necessarily mean better detection; it can also reflect
barcode inflation or spurious alignment in that pipeline.

```{r marker-count-plot, fig.width=ng_fig_size(length(markers), 1)$w, fig.height=ng_fig_size(length(markers), 1)$h}
expr_ok = marker_expr[!is.na(n_expressing)]

ggplot(expr_ok, aes(x = marker, y = n_expressing, fill = pipeline)) +
    geom_col(position = "dodge") +
    scale_fill_manual(values = cbbPalette) +
    theme_ng() +
    theme(aspect.ratio = NULL) +
    labs(x = NULL, y = "cells expressing (count > 0)", fill = NULL)
```

Fraction of recovered cells that are positive for each marker. Dividing by total
cell count removes the effect of different pipeline yields, so pipelines can be
compared on the proportion of their captured population that is marker-positive.

```{r marker-pct-plot, fig.width=ng_fig_size(length(markers), 1)$w, fig.height=ng_fig_size(length(markers), 1)$h}
ggplot(expr_ok, aes(x = marker, y = pct_expressing, fill = pipeline)) +
    geom_col(position = "dodge") +
    scale_fill_manual(values = cbbPalette) +
    theme_ng() +
    theme(aspect.ratio = NULL) +
    labs(x = NULL, y = "percent expressing (%)", fill = NULL)
```

### Dotplot summary

Dot size encodes the fraction of cells expressing each marker; colour encodes
mean log-normalised expression in those cells. Together they distinguish markers
that are rare but strongly expressed from those that are broadly detected at low
levels. Rows are pipelines; columns are markers.

```{r dotplot, fig.width=ng_fig_size(length(markers), n_pipelines)$w, fig.height=ng_fig_size(length(markers), n_pipelines)$h}
expr_ok[, marker_label := paste0(marker, "\n", cell_type)]

ggplot(expr_ok,
       aes(x = marker_label, y = pipeline,
           size = pct_expressing,
           colour = mean_lognorm_expr)) +
    geom_point() +
    scale_colour_viridis_c(option = "magma", direction = 1, na.value = "grey90") +
    scale_size_continuous(range = c(0.5, 6)) +
    theme_ng() +
    theme(aspect.ratio = NULL) +
    labs(x = NULL, y = NULL,
         size = "% expressing",
         colour = "mean log norm")
```

## Session info

```{r session-info}
sessionInfo()
```
