---
title: "rhapsodist: biology report"
author: "Izaskun Mallona"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    self_contained: true
params:
  working_dir: "/tmp/rhapsodist_simulate"
  sample: "simulated"
  aligners: "starsolo,kallisto,alevin"
  has_sbg: false
  markers_file: ""
  guide_csv: ""
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
    error = TRUE,
    echo = TRUE,
    cache = FALSE,
    dev = "png",
    dev.args = list(type = "cairo-png"),
    warning = FALSE,
    message = FALSE
)

suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(Seurat)
    library(scuttle)
    library(ggplot2)
    library(patchwork)
    library(data.table)
    library(Matrix)
    library(viridis)
    library(UpSetR)
})

source(file.path(dirname(knitr::current_input()), "..", "ggtheme.R"))

wd = params$working_dir
samp = params$sample
aligners = strsplit(params$aligners, ",")[[1]]
has_sbg = isTRUE(params$has_sbg) || identical(params$has_sbg, "true")

marker_meta = read.table(params$markers_file, header = TRUE, sep = "\t",
                         stringsAsFactors = FALSE)
markers = marker_meta$marker

if (has_sbg) aligners = c(aligners, "sbg")
```

## Data loading

Each pipeline produces a SingleCellExperiment saved as an RDS file. Gene symbols
sit in rowData under different column names depending on the aligner (starsolo:
`name`, kallisto: `SYMBOL`, alevin: Ensembl IDs as rownames with no symbol
column). SCE rownames are renamed to gene symbols before building Seurat objects
so that marker lookups work the same way across pipelines.

```{r load-sces}
load_sce = function(aligner, wd, samp) {
    fn = file.path(wd, aligner, samp, paste0(samp, "_", aligner, "_sce.rds"))
    if (!file.exists(fn)) {
        warning("SCE not found: ", fn)
        return(NULL)
    }
    readRDS(fn)
}

sces = setNames(lapply(aligners, load_sce, wd = wd, samp = samp), aligners)
sces = Filter(Negate(is.null), sces)

rm_zeros = function(sce) {
    sce = sce[rowSums(counts(sce)) > 0, ]
    sce[, colSums(counts(sce)) > 0]
}
sces = lapply(sces, rm_zeros)

symbol_rownames = function(sce) {
    rd = rowData(sce)
    sym = rownames(sce)
    for (col in c("name", "SYMBOL", "gene_name")) {
        if (col %in% colnames(rd)) {
            sym = as.character(rd[[col]])
            break
        }
    }
    rownames(sce) = make.unique(sym)
    sce
}

sces = lapply(sces, symbol_rownames)

cat("Pipelines loaded:", paste(names(sces), collapse = ", "), "\n")
sapply(sces, dim) |> `rownames<-`(c("n_genes", "n_cells"))
```

## QC summary

Cell and library quality metrics per pipeline. Differences in cell count come
from each aligner's barcode calling approach. Median UMI and genes detected show
how many counts each aligner assigns per cell.

```{r qc-table}
qc_df = rbindlist(lapply(names(sces), function(m) {
    sce = addPerCellQCMetrics(sces[[m]],
        subsets = list(mito = grep("^mt-|^MT-", rownames(sces[[m]]), value = TRUE)))
    data.table(
        pipeline = m,
        n_cells = ncol(sce),
        n_genes_total = nrow(sce),
        median_umi = round(median(sce$sum)),
        median_genes = round(median(sce$detected)),
        median_pct_mito = round(median(sce$subsets_mito_percent), 2)
    )
}))

saveRDS(qc_df, file.path(wd, paste0(samp, "_biology_qc.rds")))
knitr::kable(qc_df, caption = "Per-pipeline QC summary.")
```

Cells per pipeline, median UMI per cell, and median genes detected per cell.
Bars allow quick visual comparison; exact values are in the table above.

```{r qc-barplots, fig.width=ng_fig_size(3, 1)$w, fig.height=ng_fig_size(3, 1)$h}
qc_long = melt(
    qc_df[, .(pipeline, n_cells, median_umi, median_genes)],
    id.vars = "pipeline", variable.name = "metric", value.name = "value"
)

ggplot(qc_long, aes(x = pipeline, y = value, fill = pipeline)) +
    geom_col() +
    geom_text(aes(label = value), vjust = -0.4, size = 2.5) +
    facet_wrap(~metric, scales = "free_y") +
    scale_fill_manual(values = cbbPalette) +
    theme_ng() +
    theme(aspect.ratio = NULL, legend.position = "none") +
    labs(x = NULL, y = NULL)
```

Per-cell UMI and gene count distributions across pipelines. A shift to the right
means more counts per cell. Multimodal distributions may come from doublets or
cell populations of very different sizes.

```{r umi-gene-density, fig.width=ng_fig_size(2, 1)$w, fig.height=ng_fig_size(2, 1)$h}
umi_df = rbindlist(lapply(names(sces), function(m) {
    data.table(pipeline = m, umi = colSums(counts(sces[[m]])))
}))

gene_df = rbindlist(lapply(names(sces), function(m) {
    data.table(pipeline = m, genes = colSums(counts(sces[[m]]) > 0))
}))

p_umi = ggplot(umi_df, aes(x = log10(umi + 1), colour = pipeline)) +
    geom_density(linewidth = 0.6) +
    scale_colour_manual(values = cbbPalette) +
    theme_ng() +
    theme(aspect.ratio = NULL) +
    labs(x = "log10(UMI + 1)", y = "density", colour = NULL)

p_gene = ggplot(gene_df, aes(x = log10(genes + 1), colour = pipeline)) +
    geom_density(linewidth = 0.6) +
    scale_colour_manual(values = cbbPalette) +
    theme_ng() +
    theme(aspect.ratio = NULL) +
    labs(x = "log10(genes + 1)", y = "density", colour = NULL)

p_umi + p_gene + plot_layout(guides = "collect")
```

## Per-pipeline clustering and marker expression {.tabset}

For each pipeline: 2000 highly variable genes by VST, 30 PCs, Louvain clustering
at resolution 0.5, UMAP on the top 20 PCs. Feature plots show log-normalised
expression; cells with zero counts are shown in grey.

```{r build-seurat-objects}
make_seurat = function(sce) {
    mat = as(counts(sce), "CsparseMatrix")
    so = CreateSeuratObject(counts = mat, project = mainExpName(sce),
                            min.cells = 0, min.features = 0)
    so = NormalizeData(so, normalization.method = "LogNormalize",
                       scale.factor = 1e4, verbose = FALSE)
    so = FindVariableFeatures(so, selection.method = "vst", nfeatures = 2000,
                              verbose = FALSE)
    so = ScaleData(so, verbose = FALSE)
    so = RunPCA(so, features = VariableFeatures(so), npcs = 30, verbose = FALSE)
    so = FindNeighbors(so, dims = 1:20, verbose = FALSE)
    so = FindClusters(so, resolution = 0.5, verbose = FALSE)
    so = RunUMAP(so, dims = 1:20, verbose = FALSE)
    so
}

seurat_objects = list()
for (pipe in names(sces)) {
    cache_fn = file.path(wd, paste0(samp, "_biology_", pipe, "_seurat.rds"))
    if (file.exists(cache_fn)) {
        cat(pipe, ": loading cached Seurat object\n")
        seurat_objects[[pipe]] = readRDS(cache_fn)
    } else {
        cat(pipe, ": building Seurat object\n")
        seurat_objects[[pipe]] = make_seurat(sces[[pipe]])
        saveRDS(seurat_objects[[pipe]], cache_fn)
    }
}

rm(sces)
gc()
```

```{r per-pipeline-tabs, results='asis', fig.width=ng_fig_size(4, 4)$w, fig.height=ng_fig_size(4, 4)$h}
for (pipe in names(seurat_objects)) {
    so = seurat_objects[[pipe]]
    n_clusters = length(unique(Idents(so)))
    cluster_colours = rep(cbbPalette, ceiling(n_clusters / length(cbbPalette)))

    cat("### ", pipe, " {.tabset}\n\n")

    cat("#### UMAP\n\n")
    cat("Each dot is a cell; colour encodes Louvain cluster. Nearby cells are",
        "transcriptionally similar. The number and separation of clusters shows",
        "how many cell types this pipeline detects.\n\n")
    print(
        DimPlot(so, reduction = "umap", label = TRUE, label.size = 2.5,
                cols = cluster_colours[seq_len(n_clusters)]) +
        theme_ng() +
        labs(title = pipe, colour = "cluster")
    )
    cat("\n\n")

    cat("#### PCA\n\n")
    cat("PC1 vs PC2 coloured by cluster. Compare with the UMAP to check whether",
        "the top PCs separate the same groups.\n\n")
    print(
        DimPlot(so, reduction = "pca", label = TRUE, label.size = 2.5,
                cols = cluster_colours[seq_len(n_clusters)]) +
        theme_ng() +
        labs(title = pipe, colour = "cluster")
    )
    cat("\n\n")

    for (i in seq_len(nrow(marker_meta))) {
        marker = marker_meta$marker[i]
        cell_type = marker_meta$cell_type[i]
        cat("####", marker, "\n\n")
        if (marker %in% rownames(so)) {
            cat("Log-normalised expression of", marker,
                paste0("(", cell_type, ")"),
                "on the", pipe, "UMAP.",
                "Brighter cells carry higher expression;",
                "grey cells have zero counts for this gene.\n\n")
            print(
                FeaturePlot(so, reduction = "umap", features = marker,
                            order = TRUE, pt.size = 0.4) +
                scale_colour_viridis_c(option = "magma", direction = 1,
                                       na.value = "grey90") +
                theme_ng() +
                labs(title = paste0(marker, " (", cell_type, ")"),
                     subtitle = pipe,
                     colour = "log norm")
            )
        } else {
            cat(marker, "not detected in the", pipe, "feature set.\n")
        }
        cat("\n\n")
    }
}
```

## Guide RNA assignments

This dataset is from a pooled CRISPR screen. The authors provide per-cell guide
assignments from their own processing of a separate amplicon library. We do not
quantify guides here; we use that table only as cell metadata. Barcodes in the
table use CB1_CB2_CB3 format (three 9-bp segments separated by underscores); we
remove underscores to match the 27-bp concatenated format this pipeline produces.
For cells with more than one detected guide we take the one with the highest UMI
count. The join rate is the fraction of cells from each aligner that match a
guide assignment; it shows how well each aligner's barcodes agree with the
original study.

```{r guide-assignments, results='asis', fig.width=ng_fig_size(length(seurat_objects), 1)$w, fig.height=ng_fig_size(length(seurat_objects), 1)$h}
guide_csv = params$guide_csv

if (nchar(guide_csv) > 0 && file.exists(guide_csv)) {
    guides = fread(guide_csv)
    guides[, barcode_clean := gsub("_", "", barcode)]
    setorder(guides, barcode_clean, -counts)
    guides_top = guides[, .SD[1L], by = barcode_clean]
    guides_top[, guide_gene := sub("_sgRNA.*", "", sgRNA)]

    join_stats = rbindlist(lapply(names(seurat_objects), function(pipe) {
        cells = colnames(seurat_objects[[pipe]])
        n_matched = sum(cells %in% guides_top$barcode_clean)
        data.table(
            pipeline = pipe,
            n_cells = length(cells),
            n_matched = n_matched,
            join_rate_pct = round(100 * n_matched / length(cells), 1)
        )
    }))

    saveRDS(join_stats, file.path(wd, paste0(samp, "_biology_guide_join.rds")))
    print(knitr::kable(join_stats,
        caption = paste("Fraction of pipeline cells matched to an author-assigned guide.",
                        "A higher join rate indicates better barcode concordance",
                        "with the original study's processing.")))

    print(
        ggplot(join_stats, aes(x = pipeline, y = join_rate_pct, fill = pipeline)) +
            geom_col() +
            geom_text(aes(label = paste0(join_rate_pct, "%")), vjust = -0.4, size = 3) +
            scale_fill_manual(values = cbbPalette) +
            scale_y_continuous(limits = c(0, 105)) +
            theme_ng() +
            theme(aspect.ratio = NULL, legend.position = "none") +
            labs(x = NULL, y = "cells with guide assignment (%)",
                 title = "Guide barcode join rate per pipeline")
    )

    top_guides = guides_top[, .N, by = guide_gene][order(-N)][seq_len(min(20L, .N))]

    print(
        ggplot(top_guides, aes(x = reorder(guide_gene, N), y = N)) +
            geom_col(fill = cbbPalette[1]) +
            coord_flip() +
            theme_ng() +
            theme(aspect.ratio = NULL) +
            labs(x = NULL, y = "cells assigned",
                 title = "Top 20 guide targets (all barcodes)")
    )

    for (pipe in names(seurat_objects)) {
        cells = colnames(seurat_objects[[pipe]])
        seurat_objects[[pipe]]$guide_assigned = ifelse(
            cells %in% guides_top$barcode_clean, "assigned", "unassigned")
    }
} else {
    cat("No guide assignment file provided for this sample.\n")
}
```

```{r guide-umap, results='asis', fig.width=ng_fig_size(3, 3)$w, fig.height=ng_fig_size(3, 3)$h}
if (nchar(guide_csv) > 0 && file.exists(guide_csv) &&
    "guide_assigned" %in% colnames(seurat_objects[[1]]@meta.data)) {

    cat("### Guide assignment on UMAP {.tabset}\n\n")
    cat("Each cell is coloured by whether its barcode matched a guide assignment.",
        "Unassigned cells either had no guide detected or their barcode did not",
        "match the author-provided table.\n\n")

    for (pipe in names(seurat_objects)) {
        cat("####", pipe, "\n\n")
        print(
            DimPlot(seurat_objects[[pipe]], reduction = "umap",
                    group.by = "guide_assigned", pt.size = 0.3,
                    cols = c(assigned = cbbPalette[2], unassigned = "grey85")) +
                theme_ng() +
                labs(title = pipe, colour = NULL)
        )
        cat("\n\n")
    }
}
```

## Marker expression per pipeline

A cell is counted as expressing a marker if its raw count exceeds zero. Mean
expression is computed only over expressing cells using log-normalised values.

```{r marker-expression-table}
marker_expr = rbindlist(lapply(names(seurat_objects), function(pipe) {
    so = seurat_objects[[pipe]]
    n_total = ncol(so)
    rbindlist(lapply(seq_len(nrow(marker_meta)), function(i) {
        marker = marker_meta$marker[i]
        cell_type = marker_meta$cell_type[i]
        if (!(marker %in% rownames(so))) {
            return(data.table(
                pipeline = pipe,
                marker = marker,
                cell_type = cell_type,
                n_expressing = NA_integer_,
                pct_expressing = NA_real_,
                mean_lognorm_expr = NA_real_,
                n_total_cells = n_total
            ))
        }
        raw = GetAssayData(so, layer = "counts")[marker, ]
        norm = GetAssayData(so, layer = "data")[marker, ]
        n_pos = sum(raw > 0)
        data.table(
            pipeline = pipe,
            marker = marker,
            cell_type = cell_type,
            n_expressing = n_pos,
            pct_expressing = round(100 * n_pos / n_total, 2),
            mean_lognorm_expr = round(if (n_pos > 0) mean(norm[raw > 0]) else NA_real_, 3),
            n_total_cells = n_total
        )
    }))
}))

saveRDS(marker_expr, file.path(wd, paste0(samp, "_biology_marker_expr.rds")))
knitr::kable(
    marker_expr[order(marker, pipeline)],
    caption = paste(
        "Cells positive for each marker (raw count > 0) per pipeline.",
        "Mean log-normalised expression is computed over expressing cells only."
    )
)

cb_umi_dt = rbindlist(lapply(names(seurat_objects), function(pipe) {
    data.table(pipeline = pipe,
               barcode = colnames(seurat_objects[[pipe]]),
               total_umi = seurat_objects[[pipe]]$nCount_RNA)
}))
saveRDS(cb_umi_dt, file.path(wd, paste0(samp, "_biology_cb_umi.rds")))

n_pipelines = length(seurat_objects)
rm(seurat_objects)
gc()
```

## Cross-pipeline comparison

### Pseudobulk count deviation

For each pipeline, gene counts are summed across all cells to produce a
pseudobulk vector. Pairwise mean absolute relative difference (MARD) between
pipelines is shown as a heatmap. MARD = mean(|x - y| / ((x + y) / 2)) over
genes where at least one pipeline has non-zero counts. Lower values mean closer
agreement.

```{r pseudobulk-mard, fig.width=ng_fig_size(n_pipelines, n_pipelines)$w, fig.height=ng_fig_size(n_pipelines, n_pipelines)$h}
sces_pb = lapply(aligners, function(aligner) {
    fn = file.path(wd, aligner, samp, paste0(samp, "_", aligner, "_sce.rds"))
    if (!file.exists(fn)) return(NULL)
    sce = readRDS(fn)
    sce = sce[rowSums(counts(sce)) > 0, colSums(counts(sce)) > 0]
    rd = rowData(sce)
    sym = rownames(sce)
    for (col in c("name", "SYMBOL", "gene_name")) {
        if (col %in% colnames(rd)) { sym = as.character(rd[[col]]); break }
    }
    rownames(sce) = make.unique(sym)
    rowSums(counts(sce))
})
names(sces_pb) = aligners
sces_pb = Filter(Negate(is.null), sces_pb)

shared_genes = Reduce(intersect, lapply(sces_pb, names))
pb_mat = do.call(cbind, lapply(sces_pb, function(v) v[shared_genes]))

mard_mat = outer(
    colnames(pb_mat), colnames(pb_mat),
    Vectorize(function(a, b) {
        x = pb_mat[, a]
        y = pb_mat[, b]
        denom = (x + y) / 2
        mean(abs(x - y)[denom > 0] / denom[denom > 0], na.rm = TRUE)
    })
)
dimnames(mard_mat) = list(colnames(pb_mat), colnames(pb_mat))
mard_long = as.data.frame(as.table(mard_mat))
names(mard_long) = c("pipeline1", "pipeline2", "mard")

ggplot(mard_long, aes(pipeline1, pipeline2, fill = mard * 100)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = sprintf("%.2f%%", mard * 100)), size = 3.2) +
    scale_fill_distiller(palette = "RdYlGn", direction = -1, limits = c(0, NA)) +
    theme_ng() +
    theme(aspect.ratio = NULL,
          axis.text.x = element_text(angle = 30, hjust = 1)) +
    labs(title = "Pairwise pseudobulk MARD",
         subtitle = "Lower is better",
         x = NULL, y = NULL, fill = "MARD (%)")
```

### Barcode overlap

How many cell barcodes are shared across pipelines. Bars unique to one aligner
show barcodes that only that aligner called. Intersections show barcodes called
by more than one aligner.

```{r barcode-overlap, fig.width=8, fig.height=5}
bc_lists = split(cb_umi_dt$barcode, cb_umi_dt$pipeline)
upset(fromList(bc_lists), order.by = "freq", text.scale = 1.4,
      point.size = 3, line.size = 0.8, mb.ratio = c(0.6, 0.4))
```

### Per-barcode UMI concordance

For each pair of pipelines, only barcodes detected by both are compared.
Mean UMI (x-axis) and difference in UMI (y-axis) follow the Bland-Altman
convention. Dashed lines show mean bias and 1.96 SD limits of agreement.

```{r bland-altman-umi, fig.width=ng_fig_size(3, 3)$w, fig.height=ng_fig_size(3, 3)$h}
pipes = unique(cb_umi_dt$pipeline)
pairs = combn(pipes, 2, simplify = FALSE)

ba_plots = lapply(pairs, function(pair) {
    d1 = cb_umi_dt[pipeline == pair[1], .(barcode, umi1 = total_umi)]
    d2 = cb_umi_dt[pipeline == pair[2], .(barcode, umi2 = total_umi)]
    shared = merge(d1, d2, by = "barcode")
    shared[, mean_umi := (log10(umi1 + 1) + log10(umi2 + 1)) / 2]
    shared[, diff_umi := log10(umi1 + 1) - log10(umi2 + 1)]
    bias = mean(shared$diff_umi)
    loa  = sd(shared$diff_umi) * 1.96
    ggplot(shared, aes(x = mean_umi, y = diff_umi)) +
        geom_point(alpha = 0.1, size = 0.3) +
        geom_hline(yintercept = bias, colour = cbbPalette[2], linewidth = 0.6) +
        geom_hline(yintercept = c(bias - loa, bias + loa),
                   colour = cbbPalette[2], linewidth = 0.4, linetype = "dashed") +
        theme_ng() +
        theme(aspect.ratio = NULL) +
        labs(x = "mean log10(UMI + 1)",
             y = "diff log10(UMI + 1)",
             title = paste(pair[1], "vs", pair[2]),
             subtitle = sprintf("n = %d shared barcodes, bias = %.3f", nrow(shared), bias))
})

wrap_plots(ba_plots)
```

### Cell recovery

Total cells called by each pipeline. Differences come from each aligner's
barcode calling and filtering, not from biology; these numbers are the denominator
for the percentage calculations below.

```{r cell-recovery, fig.width=ng_fig_size(n_pipelines, 1)$w, fig.height=ng_fig_size(n_pipelines, 1)$h}
ncell_df = unique(marker_expr[, .(pipeline, n_total_cells)])

ggplot(ncell_df, aes(x = pipeline, y = n_total_cells, fill = pipeline)) +
    geom_col() +
    geom_text(aes(label = n_total_cells), vjust = -0.4, size = 3) +
    scale_fill_manual(values = cbbPalette) +
    theme_ng() +
    theme(aspect.ratio = NULL, legend.position = "none") +
    labs(x = NULL, y = "cells recovered")
```

### Marker counts

Number of cells with at least one UMI for each marker, split by pipeline. A
higher count does not always mean better detection; it can also come from more
cells being called or spurious alignment.

```{r marker-count-plot, fig.width=ng_fig_size(length(markers), 1)$w, fig.height=ng_fig_size(length(markers), 1)$h}
expr_ok = marker_expr[!is.na(n_expressing)]

ggplot(expr_ok, aes(x = marker, y = n_expressing, fill = pipeline)) +
    geom_col(position = "dodge") +
    scale_fill_manual(values = cbbPalette) +
    theme_ng() +
    theme(aspect.ratio = NULL) +
    labs(x = NULL, y = "cells expressing (count > 0)", fill = NULL)
```

Fraction of cells that are positive for each marker. Dividing by cell count
removes the effect of each pipeline calling different numbers of cells, so the
pipelines can be compared on the same scale.

```{r marker-pct-plot, fig.width=ng_fig_size(length(markers), 1)$w, fig.height=ng_fig_size(length(markers), 1)$h}
ggplot(expr_ok, aes(x = marker, y = pct_expressing, fill = pipeline)) +
    geom_col(position = "dodge") +
    scale_fill_manual(values = cbbPalette) +
    theme_ng() +
    theme(aspect.ratio = NULL) +
    labs(x = NULL, y = "percent expressing (%)", fill = NULL)
```

### Dotplot summary

Dot size: fraction of cells with at least one count for the marker. Colour: mean
log-normalised expression in those cells. A small bright dot means the marker is
rare but highly expressed; a large pale dot means it is widespread but lowly
expressed. Rows are pipelines; columns are markers.

```{r dotplot, fig.width=ng_fig_size(length(markers), n_pipelines)$w, fig.height=ng_fig_size(length(markers), n_pipelines)$h}
expr_ok[, marker_label := paste0(marker, "\n", cell_type)]

ggplot(expr_ok,
       aes(x = marker_label, y = pipeline,
           size = pct_expressing,
           colour = mean_lognorm_expr)) +
    geom_point() +
    scale_colour_viridis_c(option = "magma", direction = 1, na.value = "grey90") +
    scale_size_continuous(range = c(0.5, 6)) +
    theme_ng() +
    theme(aspect.ratio = NULL) +
    labs(x = NULL, y = NULL,
         size = "% expressing",
         colour = "mean log norm")
```

## Session info

```{r session-info}
sessionInfo()
```
