---
title: "rhapsodist: linker QC"
author: "Izaskun Mallona"
output:
  html_document:
    toc: false
    self_contained: true
params:
  sample: "sample"
  tsv:    ""
---

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
    library(data.table)
    library(ggplot2)
})

samp <- params$sample
tsv  <- params$tsv

raw   <- readLines(tsv)
meta  <- raw[grepl("^#", raw)]
body  <- raw[!grepl("^#", raw)]
dt    <- fread(text = paste(body, collapse = "\n"))

parse_meta <- function(meta, key) {
    line <- grep(paste0("^# ", key, "\\t"), meta, value = TRUE)
    if (length(line) == 0) return(NA_character_)
    strsplit(line[1], "\t")[[1]][2]
}
detected_class <- parse_meta(meta, "detected_class")
n_reads_scanned <- as.integer(parse_meta(meta, "n_reads"))
```

## Sample: `r samp`

Auto-detected bead class: **`r detected_class`**.
Reads scanned: `r format(n_reads_scanned, big.mark = ",")`.

Each read's R1 was compared to the expected fixed linker sequences for both the
`v1` and `enhanced` chemistries. The histogram below is the total hamming
distance across the two linker regions per read (for enhanced the minimum
across the 0-3 bp diversity-insert stagger is taken). Zero means an exact
match; one means a single substitution somewhere in the fixed linker bases;
two or more means the barcode call is likely unreliable for cutadapt with
`-e 0`.

```{r plot, fig.width=7, fig.height=4.5}
if (nrow(dt) == 0) {
    cat("No reads scanned; skipping plot.\n")
} else {
    max_err_to_plot <- min(max(dt$n_errors), 10)
    plot_dt <- dt[n_errors <= max_err_to_plot]
    ggplot(plot_dt, aes(x = factor(n_errors), y = frac, fill = class)) +
        geom_col(position = position_dodge(width = 0.9)) +
        geom_text(aes(label = scales::percent(frac, accuracy = 0.1)),
                  position = position_dodge(width = 0.9),
                  vjust = -0.3, size = 2.8) +
        scale_fill_brewer(palette = "Set2") +
        theme_bw() +
        labs(title = paste0("Linker mismatches per read (", samp, ")"),
             subtitle = "Hamming distance of R1 fixed linker regions to expected sequence",
             x = "Linker mismatches per read (sum across both linkers)",
             y = "Fraction of reads", fill = "Candidate class")
}
```

## Cumulative fractions by tolerance

Each row tells you the fraction of reads that would survive a cutadapt
`-e N` trim at the linker level. Use this to pick `cb_umi_max_errors`.

```{r table}
if (nrow(dt) > 0) {
    cum_tbl <- dt[, {
        ord <- order(n_errors)
        cf  <- cumsum(frac[ord])
        .(n_errors = n_errors[ord], cumulative_frac = cf)
    }, by = class]
    out <- dcast(cum_tbl, n_errors ~ class, value.var = "cumulative_frac", fill = NA_real_)
    out <- out[n_errors <= 5]
    knitr::kable(out, digits = 4,
                 caption = "Fraction of reads within N linker mismatches, per candidate class")
}
```
