---
title: "rhapsodist: overview"

format:
  html:
    toc: true
    code-fold: true
    code-tools: true
    toc_float: true

author: Jiayi Wang
---

## Dependencies
```{r packages}
#setwd("~/Desktop/sher/rhapsodist/")
suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(Matrix)
    library(AnnotationDbi)
    library(org.Hs.eg.db)
    library(EnsDb.Hsapiens.v86)
    library(UpSetR)
    library(ggplot2)
    library(scater)
    library(scDblFinder)
    library(data.table)
    source("~/rhapsodist/scripts/index2barcode.R")
})
```


## Data loading

### rhapsodist
```{r rhapsodist}
path <- "~/rhapsodist/testrun/"
sce_alevin <- readRDS(paste0(path, 
                             "alevin/example_full/example_full_alevin_sce.rds"))
sce_kallisto <- readRDS(paste0(path,"kallisto/example_full/example_full_kallisto_sce.rds"))
sce_starsolo <- readRDS(paste0(path,"starsolo/example_full/example_full_starsolo_sce.rds"))
```

### SBG
```{r sbg}
path1 <- "~/rhapsodist/docker/testrun/unfiltered_MEX_output/"
feature <- paste0(path1, "features.tsv.gz")
count <- paste0(path1, "matrix.mtx.gz")
barcode <- paste0(path1, "barcodes.tsv.gz")
counts <- readMM(count)
features <- read.table(feature, header = FALSE, sep = "\t", 
                       stringsAsFactors = FALSE)
barcodes <- read.table(barcode, header = FALSE, sep = "\t",
                       stringsAsFactors = FALSE)

rownames(counts) <- features$V1  
colnames(counts) <- barcodes$V1 

sce_sbg <- SingleCellExperiment(assays = list(counts = counts))
colnames(sce_sbg) <- sapply(colnames(sce_sbg), index_to_sequence, 
                            bead_version="Enh")
```

### Remove cells or features without any counts
```{r}
.rm <- \(x) {
    x <- x[rowSums(counts(x))!=0,]
    x <- x[,colSums(counts(x))!=0]
    x
}
sces <- list(alevin=sce_alevin, kallisto=sce_kallisto, 
             starsolo=sce_starsolo, SBG=sce_sbg)
sces <- lapply(sces, .rm)

```

### Convert Ensembl ID to gene symbol
The number of genes and the number of cells for each method after removed cells/genes with 0 count.
```{r}
convert_ensembl_to_symbol <- \(x) {
    ensembl_ids <- rownames(x)
    ensembl_ids <- sub("\\..*$", "", ensembl_ids)
    
    gene_symbols <- mapIds(EnsDb.Hsapiens.v86,
                           keys = ensembl_ids,
                           column = "SYMBOL",
                           keytype = "GENEID",
                           multiVals = "first")
    gene_symbols[is.na(gene_symbols)] <- ensembl_ids[is.na(gene_symbols)]
    
    rowData(x) <- DataFrame(ENSEMBL = ensembl_ids, SYMBOL = gene_symbols)
    rownames(x) <- gene_symbols
    x
}
sces[1:3] <- lapply(sces[1:3], convert_ensembl_to_symbol)
mx <- sapply(sces, dim)
rownames(mx) <- c("nGene", "nCell")
mx
```



## Comparison before QC

### Cell barcodes overlapping
```{r}
barcodes <- lapply(sces, colnames)
upset(fromList(barcodes), 
      order.by = "freq",
      text.scale = 1.5)
```


### Expressed genes overlapping
```{r}
genes <- lapply(sces, rownames)
upset(fromList(genes), 
      order.by = "freq",
      text.scale = 1.5)
```

### Distribution of reads per cell
```{r, fig.width=8, fig.height=8}
reads <- lapply(sces, \(x) {
    data.table(CB=colnames(x), nReads=colSums(counts(x)), nGenes=colSums(counts(x) > 0))
}) 
reads <- rbindlist(reads, idcol = "workflow")
ggplot(reads, aes(x = log10(nReads), color = workflow)) +
    geom_density() +
    labs(
        title = "Reads per Cell Distribution by Workflow",
        x = "Reads per Cell (Log scale)",
        y = "Density",
        color = "Workflow") +
    theme_bw() +
    facet_wrap(~workflow, scales="free", ncol=2) +
    scale_color_brewer(palette = "Paired")
```

NOTE: so many cells only contain 1 count in kallisto. 
Set threshold to 1 for kallisto:
```{r}
# NOTE: so many cells only contain 1 count in kallisto
sum(colSums(counts(sces$kallisto))==1)
x <- sces$kallisto
x <- x[,colSums(counts(x))>1]
dim(x)
df <- data.table(CB=colnames(x), 
                 nReads=colSums(counts(x)), 
                 nGenes=colSums(counts(x) > 0))
ggplot(df, aes(x = log10(nReads))) +
    geom_density() +
    labs(
        title = "Reads per Cell Distribution by Workflow",
        x = "Reads per Cell (Log scale)",
        y = "Density",
        color = "Workflow") +
    theme_bw()
sces$kallisto <- x
```

### Top 20 expressed genes


::: {.panel-tabset}

### alevin
```{r alevin}
plotHighestExprs(sces[[1]], n=20)
```

### kallisto

```{r kallisto}
plotHighestExprs(sces[[2]], n=20)
# NOTE: problem plotting the figure, so print
# gene_sums <- rowSums(assay(sces[[2]], "counts"))
# top_genes <- names(sort(gene_sums, decreasing = TRUE))[1:20]
# top_genes
```

### starsolo
```{r starsolo}
plotHighestExprs(sces[[3]], n=20)
```

### SBG
```{r sbg top}
plotHighestExprs(sces[[4]], n=20)
```

::: 

## Saving
```{r}
saveRDS(sces, "~/rhapsodist/testrun/sces.rds")
```






