---
title: "sampletag demultiplexing report"
author: "rhapsodist"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    number_sections: true
params:
  counts_path:      ""
  assignments_path: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE, error = TRUE)
suppressPackageStartupMessages({
    library(data.table)
    library(Matrix)
    library(ggplot2)
})
```

# load sampletag alignment counts

each row of the input tsv is one aligned read: cb, umi, sampletag, cigar.

```{r load}
# fread + zcat is much faster than read.table(gzfile()) for large files
counts_raw <- fread(
    cmd = paste("zcat", shQuote(params$counts_path)),
    sep = "\t", header = FALSE,
    col.names = c("cb", "umi", "sampletag", "cigar")
)
cat(sprintf("total aligned sampletag reads: %d\n", nrow(counts_raw)))
cat(sprintf("unique cbs seen: %d\n", uniqueN(counts_raw$cb)))
cat(sprintf("sampletags detected: %s\n", paste(unique(counts_raw$sampletag), collapse = ", ")))
```

# aggregate per cell x sampletag

```{r aggregate}
# data.table group-by is hash-based and handles 100M+ rows fast.
# sparseMatrix avoids building a full dense table at this stage.
agg <- counts_raw[, .N, by = .(cb, sampletag)]
cb_f  <- factor(agg$cb)
tag_f <- factor(agg$sampletag)
mat <- as.matrix(sparseMatrix(
    i = as.integer(cb_f),
    j = as.integer(tag_f),
    x = agg$N,
    dims = c(nlevels(cb_f), nlevels(tag_f)),
    dimnames = list(levels(cb_f), levels(tag_f))
))

cat(sprintf("count matrix: %d cells x %d sampletags\n", nrow(mat), ncol(mat)))
```

# demultiplexing (bd rhapsody algorithm)

the bd algorithm: identify high-quality singlets (>75%), estimate per-tag
noise via a linear trend, subtract expected noise, call remaining cells
against per-tag minimum-count thresholds.

```{r demux}
# step 1: high-quality singlets
total = rowSums(mat)
props = mat / pmax(total, 1L)
dominant_idx = max.col(props, ties.method = "first")
dominant_frac = props[cbind(seq_len(nrow(props)), dominant_idx)]
dominant_tag = colnames(mat)[dominant_idx]
is_highqual = dominant_frac >= 0.75 & total > 0

# step 2: per-tag thresholds (min count among high-quality singlets for each tag)
tag_thresholds = vapply(seq_along(colnames(mat)), function(j) {
    hq = is_highqual & dominant_idx == j
    if (!any(hq)) return(NA_real_)
    min(mat[hq, j])
}, numeric(1))
names(tag_thresholds) = colnames(mat)

# step 3: noise model using high-quality singlets only
hq_counts = mat[cbind(seq_len(nrow(mat)), dominant_idx)]
noise_hq = ifelse(is_highqual, total - hq_counts, NA_real_)

# per-tag noise fraction: fraction of background attributable to each tag
noise_mat = mat
noise_mat[cbind(which(is_highqual), dominant_idx[is_highqual])] = 0L
noise_mat[!is_highqual, ] = NA_real_
tag_noise_frac = colSums(noise_mat, na.rm = TRUE)
total_noise = sum(tag_noise_frac)
if (total_noise > 0) {
    tag_noise_frac = tag_noise_frac / total_noise
} else {
    tag_noise_frac[] = 1 / ncol(mat)
}

# analytic ols slope (no intercept): noise ~ 0 + total
hq_total = total[is_highqual]
hq_noise = noise_hq[is_highqual]
slope = if (length(hq_total) >= 2 && var(hq_total) > 0) {
    max(sum(hq_total * hq_noise) / sum(hq_total^2), 0)
} else {
    0
}

# step 4: noise-corrected counts
# corrected[i, j] = mat[i, j] - slope * total[i] * tag_noise_frac[j]
predicted_noise = slope * total
corrected = mat - outer(predicted_noise, tag_noise_frac)

# step 5: call non-highqual cells against per-tag thresholds
valid = !is.na(tag_thresholds)
called_mat = matrix(FALSE, nrow(mat), ncol(mat), dimnames = dimnames(mat))
if (any(valid)) {
    called_mat[, valid] = sweep(
        corrected[, valid, drop = FALSE], 2, tag_thresholds[valid], ">"
    )
}
n_called = rowSums(called_mat)

# final classification
status = rep("undetermined", nrow(mat))
called_tag = rep(NA_character_, nrow(mat))

status[is_highqual] = "highqual"
called_tag[is_highqual] = dominant_tag[is_highqual]

is_called = !is_highqual & n_called == 1
status[is_called] = "called"
if (any(is_called)) {
    called_tag[is_called] = colnames(called_mat)[
        max.col(called_mat[is_called, , drop = FALSE], ties.method = "first")
    ]
}
status[!is_highqual & n_called > 1] = "multiplet"

demux = data.frame(
    cb = rownames(mat),
    called_tag = called_tag,
    dominant_frac = round(dominant_frac, 3),
    total_reads = total,
    predicted_noise = round(predicted_noise, 1),
    status = status,
    row.names = NULL,
    stringsAsFactors = FALSE
)

cat(sprintf("noise model slope: %.4f\n", slope))
print(table(demux$status))
print(table(demux$called_tag, useNA = "ifany"))
```

# distributions

```{r plots, fig.width=7, fig.height=4}
ggplot(demux, aes(x = total_reads, fill = status)) +
    geom_histogram(bins = 30, colour = "white") +
    labs(title = "total sampletag reads per cell", x = "reads", y = "cells") +
    theme_bw()

ggplot(demux[demux$status %in% c("highqual", "called"), ],
       aes(x = called_tag, fill = called_tag)) +
    geom_bar() +
    labs(title = "called cell counts per sampletag", x = "sampletag", y = "cells") +
    theme_bw() + theme(legend.position = "none")

ggplot(demux, aes(x = dominant_frac, fill = status)) +
    geom_histogram(bins = 30, colour = "white") +
    geom_vline(xintercept = 0.75, linetype = "dashed", colour = "red") +
    labs(title = "dominant sampletag fraction per cell",
         x = "fraction of reads from top tag", y = "cells") +
    theme_bw()

noise_plot_df = data.frame(
    total = total[is_highqual],
    noise = noise_hq[is_highqual]
)
ggplot(noise_plot_df, aes(x = total, y = noise)) +
    geom_point(size = 0.4, alpha = 0.4) +
    geom_abline(intercept = 0, slope = slope, colour = "red", linewidth = 0.8) +
    labs(title = "noise trend (high-quality singlets)",
         subtitle = sprintf("slope = %.4f", slope),
         x = "total sampletag reads", y = "noise reads") +
    theme_bw()
```

# concordance with ground truth (simulated data only)

```{r concordance}
if (!is.null(params$assignments_path) && nzchar(params$assignments_path) &&
        file.exists(params$assignments_path)) {
    truth = read.table(params$assignments_path, header = TRUE,
                       sep = "\t", stringsAsFactors = FALSE)
    colnames(truth) = c("cb", "true_tag")
    merged = merge(demux, truth, by = "cb", all.x = TRUE)
    n_concordant = sum(merged$called_tag == merged$true_tag, na.rm = TRUE)
    n_called_any = sum(merged$status %in% c("highqual", "called"))
    cat(sprintf("concordance: %d / %d called cells match ground truth (%.1f%%)\n",
                n_concordant, n_called_any,
                100 * n_concordant / max(n_called_any, 1)))
    print(table(called = merged$called_tag, truth = merged$true_tag, useNA = "ifany"))
} else {
    cat("no ground-truth assignments file provided - skipping concordance check.\n")
}
```

# session info

```{r sessioninfo}
sessionInfo()
```
